### 2023-06-15 papers/cpp2024/paper.tex
check

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall]{acmart}
\usepackage{semantic}

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2023}
\acmYear{2023}
\acmDOI{XXXXXXX.XXXXXXX}

%%
%% These commands are for a JOURNAL article.
%% \acmJournal{JACM}
%% \acmVolume{37}
%% \acmNumber{4}
%% \acmArticle{111}
%% \acmMonth{8}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
\citestyle{acmauthoryear}

\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\semV}[1]{\mathcal{V}\sem{#1}}
\newcommand{\semE}[1]{\mathcal{E}\sem{#1}}
\newcommand{\ba}{\begin{array}}
\newcommand{\ea}{\end{array}}
\newenvironment{stack}{\ba{@{}l@{}}}{\ea}
\newenvironment{branch}{\left\{\ba{@{}l@{\qquad}l@{}}}{\ea\right\}}
\newenvironment{syntax}{\[\ba{l@{\;\;}lcl}}{\ea\]}
\newcommand{\dotspace}{.\,}
\newcommand{\key}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\dyn}{\star}
\newcommand{\Dyn}{\ensuremath{\dyn}}
\newcommand{\Int}{\key{int}}
\newcommand{\Bool}{\key{bool}}
\newcommand{\lam}[1]{\lambda #1 \dotspace}
\newcommand{\app}{\;}
\newcommand{\reduce}{\longrightarrow}
\newcommand{\inj}[2]{#1 \,!\, #2}
\newcommand{\proj}[2]{#1 \,?\, #2}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{A Step-Indexed Logic in Agda}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Jeremy G. Siek}
%\authornote{Both authors contributed equally to this research.}
\email{jsiek@indiana.edu}
\orcid{0000-0002-9894-4856}
%% \author{G.K.M. Tobin}
%% \authornotemark[1]
%% \email{webmaster@marysville-ohio.com}
\affiliation{%
  \institution{Indiana University}
  \streetaddress{Luddy Hall, 700 N. Woodlawn Avenue}
  \city{Bloomington}
  \state{IN}
  \country{USA}
  \postcode{47408}
}

%% \author{Lars Th{\o}rv{\"a}ld}
%% \affiliation{%
%%   \institution{The Th{\o}rv{\"a}ld Group}
%%   \streetaddress{1 Th{\o}rv{\"a}ld Circle}
%%   \city{Hekla}
%%   \country{Iceland}}
%% \email{larst@affiliation.org}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
%\renewcommand{\shortauthors}{Trovato et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  Step-indexed logical relations are an important tool in the
  metatheory of programming
  languages~\citep{Appel:2001aa,Ahmed:2006aa}, and their application
  can be streamlined by use of modal logics such as
  LSLR~\citep{Dreyer:2011wl}.  The Iris~\citep{JUNG:2018aa} system
  provides support for reasoning about step-indexed logical relations
  in the Coq proof assistant, but there is not yet analogous support
  in the Agda proof assistant. This paper describes how to embed a
  step-indexed logic in Agda and application of the logic to a
  semantic type safety proof for an internal language for the
  Gradually-Typed Lambda Calculus.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
%\begin{CCSXML}
%\end{CCSXML}

%% \ccsdesc[500]{Computer systems organization~Embedded systems}
%% \ccsdesc[300]{Computer systems organization~Redundancy}
%% \ccsdesc{Computer systems organization~Robotics}
%% \ccsdesc[100]{Networks~Network reliability}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{step-indexed logical relations, modal logic}

%% \received{20 February 2007}
%% \received[revised]{12 March 2009}
%% \received[accepted]{5 June 2009}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Motivation}

Logical relations provide a powerful tool for proving properties of
programming languages, from type safety to relational parametricity,
the gradual guarantee, and noninterference.  Initially, logical
relations were applied to terminating languages such as the Simply
Typed Lambda Calculus~\citep{Tait:1967aa} and System
F~\citep{GIRARD72}, but \citet{Appel:2001aa} extended the technique to
handle languages with nontermination by introducing a step index into
the logical relation. After considerable experience with applying
step-indexed logical
relations~\citep{Ahmed:2004eu,Ahmed:2009aa,Neis:2009fk,Hur:2011aa},
\citet{Dreyer:2011wl} designed the LSLR logic to streamline the
reasoning about step indexes.

Figure~\ref{fig:CC}

\begin{figure}
  \raggedright Syntax
  \begin{syntax}
    \text{Types} & A,B,C,D &::=& \Bool \mid A \to B \mid \dyn \\
    \text{Ground Types} & G,H & ::= & \mathtt{bool} \mid \dyn \to \dyn \\
    \text{Booleans} & b & \in & \mathbb{B} \\
    \text{Terms} & L,M,N & ::= & x \mid b \mid \lam{x}N \mid L \app M
      \mid \inj{M}{G} \mid \proj{M}{H} \mid \mathtt{error} \\
      \text{Values} & V,W & ::= & \lam{x}N \mid \inj{V}{G}
  \end{syntax}
  Type System
  \begin{gather*}
    \inference{}{\Gamma \vdash x : \Gamma(x)}
    \quad
    \inference{}{\Gamma \vdash b : \Bool}
    \quad
    \inference{\Gamma, x : A \vdash N : B}
              {\Gamma \vdash \lam{x} N : A \to B}
    \quad              
    \inference{\Gamma \vdash L : A \to B & \Gamma \vdash M : A}
          {\Gamma \vdash L \app M : B} \\[2ex]
   \inference{\Gamma \vdash M : G}
          {\Gamma \vdash \inj{M}{G} : \dyn}
   \quad
   \inference{\Gamma \vdash M : \dyn}
             {\Gamma \vdash \proj{M}{H} : H}
   \quad
   \inference{}
             {\Gamma \vdash \mathtt{error} : A}
  \end{gather*}
Reduction
\begin{align*}
  (\lam{x} N)\app W & \reduce N[x:=W] \\
  \proj{\inj{V}{G}}{G} & \reduce V \\
  \proj{\inj{V}{G}}{H} & \reduce \texttt{error}
       \qquad\qquad\qquad\text{if } G \neq H 
\end{align*}

  \caption{Internal language (cast calculus) for the Gradually-Typed
    Lambda Calculus.}
  \label{fig:CC}
\end{figure}

\begin{align*}
    \semE{\_} &: \mathsf{Type} \to \mathsf{Term} \to \mathsf{Set} \\
    \semE{A} M &= (\semV{A} M \lor \mathsf{reducible}\ M \lor M=\mathtt{blame})
                \land (\forall N, (M \reduce N) \Rightarrow \rhd \semE{A} N)
\end{align*}

\bibliographystyle{ACM-Reference-Format}
\bibliography{all}

\end{document}
\endinput

### 2023-06-19 papers/cpp2024/paper.tex
check

@@
 \newcommand{\lam}[1]{\lambda #1 \dotspace}
-\newcommand{\app}{\;}
+\newcommand{\app}{\,}
 \newcommand{\reduce}{\longrightarrow}
@@
   can be streamlined by use of modal logics such as
-  LSLR~\citep{Dreyer:2011wl}.  The Iris~\citep{JUNG:2018aa} system
+  LSLR~\citep{Dreyer:2011wl}.  The Iris system~\citep{JUNG:2018aa}
   provides support for reasoning about step-indexed logical relations
   in the Coq proof assistant, but there is not yet analogous support
-  in the Agda proof assistant. This paper describes how to embed a
-  step-indexed logic in Agda and application of the logic to a
+  for the Agda proof assistant. This paper describes an embedding of
+  step-indexed logic in Agda and an application of the logic to a
   semantic type safety proof for an internal language for the
@@
 
-\section{Motivation}
+%TODO: Road Map/Outline of what's in the paper
 
-Logical relations provide a powerful tool for proving properties of
+\section{Introduction}
+
+Logical relations are a useful technique for proving properties of
 programming languages, from type safety to relational parametricity,
-the gradual guarantee, and noninterference.  Initially, logical
-relations were applied to terminating languages such as the Simply
-Typed Lambda Calculus~\citep{Tait:1967aa} and System
-F~\citep{GIRARD72}, but \citet{Appel:2001aa} extended the technique to
-handle languages with nontermination by introducing a step index into
-the logical relation. After considerable experience with applying
-step-indexed logical
+and noninterference.  Initially, logical relations were applied to
+terminating languages such as the Simply Typed Lambda
+Calculus~\citep{Tait:1967aa} and System F~\citep{GIRARD72}, but
+\citet{Appel:2001aa} extended the technique to handle languages with
+nontermination by introducing a step index into the logical
+relation. After some experience with applications of step-indexed
+logical
 relations~\citep{Ahmed:2004eu,Ahmed:2009aa,Neis:2009fk,Hur:2011aa},
 \citet{Dreyer:2011wl} designed the LSLR logic to streamline the
-reasoning about step indexes.
+reasoning about step indexes. This research program eventually
+produced Iris~\citep{JUNG:2018aa}, a system that supports a
+combination of step indexed logical relations and separation logic
+within the Coq proof
+assistant~\citep{The-Coq-Development-Team:2004kf,Huet:2016aa}.
+
+The author is interested in step-indexed logical relations for
+purposes of developing mechanized proofs in Agda of the Gradual
+Guarantee~\citep{Siek:2015ac} and Parametricity~\citep{REYNOLDS74C}
+for gradually-typed polymorphic
+languages~\citep{Ahmed:2011fk,Ahmed:2017aa,Igarashi:2017aa,New:2019ab,Labrada:2020tk}.
+To streamline such developments in Agda, we need support for a
+step-indexed logic in Agda. This paper fills that need by showing how
+to build a shallow embedding of a step-indexed logic in Agda, a system
+that we abbreviate as SIL. To evaluate the utility of SIL, we present
+a case study of proving semantic type safety of a simple cast
+calculus, i.e., an internal language for the Gradually-Typed Lambda
+Calculus (GTLC)~\citep{Siek:2006bh,Siek:2007qy}. The outcome of that
+study is that SIL streamlines the definition of logical relations and
+is helpful in many of the proofs, but there is room for improvement
+regarding the ergonomics of interactive proof development.
+Limitations in Agda's inference algorithm sometimes leads to verbosity
+in SIL proofs. It is worth noting that the authors have also proved
+the Gradual Guarantee for the GTLC using SIL and the lesssons from
+that experience confirm the findings in this paper.
+
+TODO: ROAD MAP
+
+\section{A Step-Indexed Logic (SIL)}
+
+Figure~\ref{fig:SIL}
+
+\begin{figure}
+  \raggedright
+Syntax
+  \begin{syntax}
+    \text{Propositions} & P,Q &::= & \top \mid \bot \mid P \land Q
+    \mid P \lor Q \mid P \Rightarrow Q \mid \forall X.P \mid \exists X.P
+    \mid \overline{M} \in R\\
+    \text{Relations} & R,S & ::= & r \mid \mu r. R \mid \{ \overline{x} \mid P \}
+  \end{syntax}
+Meaning of Propositions
+\begin{align*}
+  \sem{ P } \rho\; 0 &= \top \\
+  \sem{ \top } \rho(suc \app n) &= \top \\
+  \sem{ \bot } \rho (suc \app n) &= \bot \\
+  \sem{ P \land Q } \rho (suc \app n) &=
+    \sem{ P } \rho n \land \sem{ Q } \rho n \\
+  \sem{ P \lor Q } \rho (suc \app n) &=
+    \sem{ P } \rho n \lor \sem{ Q } \rho n \\
+  \sem{ P \Rightarrow Q } \rho (suc \app n) &=
+    \forall k \leq suc \app n,
+    \sem{ P } \rho k \implies \sem{ Q } \rho k \\
+  \sem{ \forall X.P} \rho \app (suc \app n) &= \forall a, \sem{P[a/X]} \rho (suc \app n)\\
+  \sem{ \exists X.P} \rho \app (suc \app n) &= \exists a, \sem{P[a/X]} \rho (suc \app n)\\
+  \sem{ \overline{M} \in R } \rho (suc \app n) &=
+    \sem{ R } \rho (suc \app n) \app \overline{M}\\
+  \sem{ \rhd P } \rho (suc \app n) &=
+    \sem{ P } \rho n 
+\end{align*}
+Meaning of Relations
+\begin{align*}
+  \sem{ R } \rho\, 0 \app \overline{M} &= \top \\
+  \sem{ r } \rho \app (suc \app n) \app \overline{M} &=
+    \rho(r) (suc \app n) \app \overline{M} \\
+  \sem{ \mu r. R } \rho \app (suc \app n)\app \overline{M} &=
+  F^{suc \app n}(\top) (suc \app n) \app \overline{M} \\
+  & \text{ where } F(X) = \sem{ R } \rho(r := X) \\
+  \sem{ \{ \overline{x} \mid P \} } \rho (suc \app n) \app \overline{M} &= 
+    \sem{ P[\overline{x} := \overline{M}] } \rho (suc \app n)
+\end{align*}
+  \caption{Step-Indexed Logic (SIL)}
+  \label{fig:SIL}
+\end{figure}
+
+\section{Case Study: Semantic Type Safety of a Cast Calculus}
 

### 2023-06-21 papers/cpp2024/paper.tex
check

@@
 \usepackage{semantic}
+\usepackage{agda}
+\usepackage{MnSymbol}
 
@@
 
+\usepackage{newunicodechar}
+\newunicodechar{λ}{\ensuremath{\mathnormal\lambda}}
+\newunicodechar{←}{\ensuremath{\mathnormal\from}}
+\newunicodechar{→}{\ensuremath{\mathnormal\to}}
+\newunicodechar{∀}{\ensuremath{\mathnormal\forall}}
+\newunicodechar{∃}{\ensuremath{\exists}}
+\newunicodechar{⊥}{\ensuremath{\mathnormal\bot}}
+\newunicodechar{⊤}{\ensuremath{\mathnormal\top}}
+\newunicodechar{∷}{\ensuremath{\mathnormal::}}
+\newunicodechar{ℕ}{\ensuremath{\mathbb N}}
+\newunicodechar{≤}{\ensuremath{\leq}}
+\newunicodechar{∸}{\ensuremath{\div}}
+\newunicodechar{′}{\ensuremath{'}}
+\newunicodechar{⇒}{\ensuremath{\Rightarrow}}
+\newunicodechar{₁}{\ensuremath{_1}}
+\newunicodechar{₂}{\ensuremath{_2}}
+\newunicodechar{Σ}{\ensuremath{\Sigma}}
+\newunicodechar{⊎}{\ensuremath{\cupdot}}
+\newunicodechar{ᵖ}{\ensuremath{^p}}
+\newunicodechar{≡}{\ensuremath{\equiv}}
+\newunicodechar{≢}{\ensuremath{\not\equiv}}
+\newunicodechar{∘}{\ensuremath{\circ}}
+\newunicodechar{ᵒ}{\ensuremath{^\circ}}
+\newunicodechar{▷}{\ensuremath{\rhd}}
+
 %%
@@
   for the Agda proof assistant. This paper describes an embedding of
-  step-indexed logic in Agda and an application of the logic to a
+  a step-indexed logic in Agda and an application of the logic to a
   semantic type safety proof for an internal language for the
@@
 calculus, i.e., an internal language for the Gradually-Typed Lambda
-Calculus (GTLC)~\citep{Siek:2006bh,Siek:2007qy}. The outcome of that
-study is that SIL streamlines the definition of logical relations and
-is helpful in many of the proofs, but there is room for improvement
-regarding the ergonomics of interactive proof development.
-Limitations in Agda's inference algorithm sometimes leads to verbosity
-in SIL proofs. It is worth noting that the authors have also proved
-the Gradual Guarantee for the GTLC using SIL and the lesssons from
-that experience confirm the findings in this paper.
+Calculus (GTLC)~\citep{Siek:2006bh,Siek:2007qy}. We find that SIL
+streamlines the definition of logical relations and is helpful in many
+of the proofs, but there is room for improvement regarding the
+ergonomics of interactive proof development using SIL.  Limitations in
+Agda's inference algorithm sometimes leads to verbosity in SIL
+proofs. It is worth noting that the authors have also proved the
+Gradual Guarantee for the GTLC using SIL and the lesssons from that
+experience confirm the findings in this paper.
 
-TODO: ROAD MAP
+%TODO: ROAD MAP
 
-\section{A Step-Indexed Logic (SIL)}
+%\section{A Step-Indexed Logic (SIL)}
 
-Figure~\ref{fig:SIL}
+%% Figure~\ref{fig:SIL}
 
-\begin{figure}
-  \raggedright
-Syntax
-  \begin{syntax}
-    \text{Propositions} & P,Q &::= & \top \mid \bot \mid P \land Q
-    \mid P \lor Q \mid P \Rightarrow Q \mid \forall X.P \mid \exists X.P
-    \mid \overline{M} \in R\\
-    \text{Relations} & R,S & ::= & r \mid \mu r. R \mid \{ \overline{x} \mid P \}
-  \end{syntax}
-Meaning of Propositions
-\begin{align*}
-  \sem{ P } \rho\; 0 &= \top \\
-  \sem{ \top } \rho(suc \app n) &= \top \\
-  \sem{ \bot } \rho (suc \app n) &= \bot \\
-  \sem{ P \land Q } \rho (suc \app n) &=
-    \sem{ P } \rho n \land \sem{ Q } \rho n \\
-  \sem{ P \lor Q } \rho (suc \app n) &=
-    \sem{ P } \rho n \lor \sem{ Q } \rho n \\
-  \sem{ P \Rightarrow Q } \rho (suc \app n) &=
-    \forall k \leq suc \app n,
-    \sem{ P } \rho k \implies \sem{ Q } \rho k \\
-  \sem{ \forall X.P} \rho \app (suc \app n) &= \forall a, \sem{P[a/X]} \rho (suc \app n)\\
-  \sem{ \exists X.P} \rho \app (suc \app n) &= \exists a, \sem{P[a/X]} \rho (suc \app n)\\
-  \sem{ \overline{M} \in R } \rho (suc \app n) &=
-    \sem{ R } \rho (suc \app n) \app \overline{M}\\
-  \sem{ \rhd P } \rho (suc \app n) &=
-    \sem{ P } \rho n 
-\end{align*}
-Meaning of Relations
-\begin{align*}
-  \sem{ R } \rho\, 0 \app \overline{M} &= \top \\
-  \sem{ r } \rho \app (suc \app n) \app \overline{M} &=
-    \rho(r) (suc \app n) \app \overline{M} \\
-  \sem{ \mu r. R } \rho \app (suc \app n)\app \overline{M} &=
-  F^{suc \app n}(\top) (suc \app n) \app \overline{M} \\
-  & \text{ where } F(X) = \sem{ R } \rho(r := X) \\
-  \sem{ \{ \overline{x} \mid P \} } \rho (suc \app n) \app \overline{M} &= 
-    \sem{ P[\overline{x} := \overline{M}] } \rho (suc \app n)
-\end{align*}
-  \caption{Step-Indexed Logic (SIL)}
-  \label{fig:SIL}
-\end{figure}
+%% \begin{figure}
+%%   \raggedright
+%% Syntax
+%%   \begin{syntax}
+%%     \text{Propositions} & P,Q &::= & \top \mid \bot \mid P \land Q
+%%     \mid P \lor Q \mid P \Rightarrow Q \mid \forall X.P \mid \exists X.P
+%%     \mid \overline{M} \in R\\
+%%     \text{Relations} & R,S & ::= & r \mid \mu r. R \mid \{ \overline{x} \mid P \}
+%%   \end{syntax}
+%% Meaning of Propositions
+%% \begin{align*}
+%%   \sem{ P } \rho\; 0 &= \top \\
+%%   \sem{ \top } \rho(suc \app n) &= \top \\
+%%   \sem{ \bot } \rho (suc \app n) &= \bot \\
+%%   \sem{ P \land Q } \rho (suc \app n) &=
+%%     \sem{ P } \rho n \land \sem{ Q } \rho n \\
+%%   \sem{ P \lor Q } \rho (suc \app n) &=
+%%     \sem{ P } \rho n \lor \sem{ Q } \rho n \\
+%%   \sem{ P \Rightarrow Q } \rho (suc \app n) &=
+%%     \forall k \leq suc \app n,
+%%     \sem{ P } \rho k \implies \sem{ Q } \rho k \\
+%%   \sem{ \forall X.P} \rho \app (suc \app n) &= \forall a, \sem{P[a/X]} \rho (suc \app n)\\
+%%   \sem{ \exists X.P} \rho \app (suc \app n) &= \exists a, \sem{P[a/X]} \rho (suc \app n)\\
+%%   \sem{ \overline{M} \in R } \rho (suc \app n) &=
+%%     \sem{ R } \rho (suc \app n) \app \overline{M}\\
+%%   \sem{ \rhd P } \rho (suc \app n) &=
+%%     \sem{ P } \rho n 
+%% \end{align*}
+%% Meaning of Relations
+%% \begin{align*}
+%%   \sem{ R } \rho\, 0 \app \overline{M} &= \top \\
+%%   \sem{ r } \rho \app (suc \app n) \app \overline{M} &=
+%%     \rho(r) (suc \app n) \app \overline{M} \\
+%%   \sem{ \mu r. R } \rho \app (suc \app n)\app \overline{M} &=
+%%   F^{suc \app n}(\top) (suc \app n) \app \overline{M} \\
+%%   & \text{ where } F(X) = \sem{ R } \rho(r := X) \\
+%%   \sem{ \{ \overline{x} \mid P \} } \rho (suc \app n) \app \overline{M} &= 
+%%     \sem{ P[\overline{x} := \overline{M}] } \rho (suc \app n)
+%% \end{align*}
+%%   \caption{Step-Indexed Logic (SIL)}
+%%   \label{fig:SIL}
+%% \end{figure}
 
-\section{Case Study: Semantic Type Safety of a Cast Calculus}
+%\section{Case Study: Semantic Type Safety of a Cast Calculus}
 
-Figure~\ref{fig:CC}
+%% Figure~\ref{fig:CC}
 
-\begin{figure}
-  \raggedright Syntax
-  \begin{syntax}
-    \text{Types} & A,B,C,D &::=& \Bool \mid A \to B \mid \dyn \\
-    \text{Ground Types} & G,H & ::= & \mathtt{bool} \mid \dyn \to \dyn \\
-    \text{Booleans} & b & \in & \mathbb{B} \\
-    \text{Terms} & L,M,N & ::= & x \mid b \mid \lam{x}N \mid L \app M
-      \mid \inj{M}{G} \mid \proj{M}{H} \mid \mathtt{error} \\
-      \text{Values} & V,W & ::= & \lam{x}N \mid \inj{V}{G}
-  \end{syntax}
-  Type System
-  \begin{gather*}
-    \inference{}{\Gamma \vdash x : \Gamma(x)}
-    \quad
-    \inference{}{\Gamma \vdash b : \Bool}
-    \quad
-    \inference{\Gamma, x : A \vdash N : B}
-              {\Gamma \vdash \lam{x} N : A \to B}
-    \quad              
-    \inference{\Gamma \vdash L : A \to B & \Gamma \vdash M : A}
-          {\Gamma \vdash L \app M : B} \\[2ex]
-   \inference{\Gamma \vdash M : G}
-          {\Gamma \vdash \inj{M}{G} : \dyn}
-   \quad
-   \inference{\Gamma \vdash M : \dyn}
-             {\Gamma \vdash \proj{M}{H} : H}
-   \quad
-   \inference{}
-             {\Gamma \vdash \mathtt{error} : A}
-  \end{gather*}
-Reduction
-\begin{align*}
-  (\lam{x} N)\app W & \reduce N[x:=W] \\
-  \proj{\inj{V}{G}}{G} & \reduce V \\
-  \proj{\inj{V}{G}}{H} & \reduce \texttt{error}
-       \qquad\qquad\qquad\text{if } G \neq H 
-\end{align*}
+%% \begin{figure}
+%%   \raggedright Syntax
+%%   \begin{syntax}
+%%     \text{Types} & A,B,C,D &::=& \Bool \mid A \to B \mid \dyn \\
+%%     \text{Ground Types} & G,H & ::= & \mathtt{bool} \mid \dyn \to \dyn \\
+%%     \text{Booleans} & b & \in & \mathbb{B} \\
+%%     \text{Terms} & L,M,N & ::= & x \mid b \mid \lam{x}N \mid L \app M
+%%       \mid \inj{M}{G} \mid \proj{M}{H} \mid \mathtt{error} \\
+%%       \text{Values} & V,W & ::= & \lam{x}N \mid \inj{V}{G}
+%%   \end{syntax}
+%%   Type System
+%%   \begin{gather*}
+%%     \inference{}{\Gamma \vdash x : \Gamma(x)}
+%%     \quad
+%%     \inference{}{\Gamma \vdash b : \Bool}
+%%     \quad
+%%     \inference{\Gamma, x : A \vdash N : B}
+%%               {\Gamma \vdash \lam{x} N : A \to B}
+%%     \quad              
+%%     \inference{\Gamma \vdash L : A \to B & \Gamma \vdash M : A}
+%%           {\Gamma \vdash L \app M : B} \\[2ex]
+%%    \inference{\Gamma \vdash M : G}
+%%           {\Gamma \vdash \inj{M}{G} : \dyn}
+%%    \quad
+%%    \inference{\Gamma \vdash M : \dyn}
+%%              {\Gamma \vdash \proj{M}{H} : H}
+%%    \quad
+%%    \inference{}
+%%              {\Gamma \vdash \mathtt{error} : A}
+%%   \end{gather*}
+%% Reduction
+%% \begin{align*}
+%%   (\lam{x} N)\app W & \reduce N[x:=W] \\
+%%   \proj{\inj{V}{G}}{G} & \reduce V \\
+%%   \proj{\inj{V}{G}}{H} & \reduce \texttt{error}
+%%        \qquad\qquad\qquad\text{if } G \neq H 
+%% \end{align*}
 
-  \caption{Internal language (cast calculus) for the Gradually-Typed
-    Lambda Calculus.}
-  \label{fig:CC}
-\end{figure}
+%%   \caption{Internal language (cast calculus) for the Gradually-Typed
+%%     Lambda Calculus.}
+%%   \label{fig:CC}
+%% \end{figure}
 
-\begin{align*}
-    \semE{\_} &: \mathsf{Type} \to \mathsf{Term} \to \mathsf{Set} \\
-    \semE{A} M &= (\semV{A} M \lor \mathsf{reducible}\ M \lor M=\mathtt{blame})
-                \land (\forall N, (M \reduce N) \Rightarrow \rhd \semE{A} N)
-\end{align*}
+%% \begin{align*}
+%%     \semE{\_} &: \mathsf{Type} \to \mathsf{Term} \to \mathsf{Set} \\
+%%     \semE{A} M &= (\semV{A} M \lor \mathsf{reducible}\ M \lor M=\mathtt{blame})
+%%                 \land (\forall N, (M \reduce N) \Rightarrow \rhd \semE{A} N)
+%% \end{align*}
+
+\input{latex/cpp2024/StepIndexedLogic.tex}
 

### 2023-06-23 papers/cpp2024/paper.tex
check

@@
 \newunicodechar{→}{\ensuremath{\mathnormal\to}}
+\newunicodechar{↓}{\ensuremath{\mathnormal\downarrow}}
 \newunicodechar{∀}{\ensuremath{\mathnormal\forall}}
@@
 \newunicodechar{▷}{\ensuremath{\rhd}}
+\newunicodechar{∈}{\ensuremath{\in}}
+\newunicodechar{⇔}{\ensuremath{\iff}}
+\newunicodechar{⩦}{\ensuremath{\equiv}} % !!
+\newunicodechar{μ}{\ensuremath{\mu}}
+\newunicodechar{∋}{\ensuremath{\ni}}
+\newunicodechar{ˢ}{\ensuremath{^s}}
+\newunicodechar{ᵈ}{\ensuremath{^d}}
+\newunicodechar{Γ}{\ensuremath{\Gamma}}
+\newunicodechar{∅}{\ensuremath{\emptyset}}
+\newunicodechar{δ}{\ensuremath{\delta}}
 
@@
   provides support for reasoning about step-indexed logical relations
-  in the Coq proof assistant, but there is not yet analogous support
-  for the Agda proof assistant. This paper describes an embedding of
-  a step-indexed logic in Agda and an application of the logic to a
-  semantic type safety proof for an internal language for the
-  Gradually-Typed Lambda Calculus.
+  in the Coq proof assistant. This paper describes a shallow embedding
+  of a step-indexed logic in Agda and its application to a semantic
+  type safety proof for an internal language of the Gradually-Typed
+  Lambda Calculus.
 \end{abstract}
@@
 
-%TODO: ROAD MAP
+TODO: ROAD MAP
 

### 2023-07-05 papers/cpp2024/paper.tex
progress
check
trying to organize
a little progress

@@
 \newcommand{\proj}[2]{#1 \,?\, #2}
+\newcommand{\kapprox}{\mathrm{approx}}
+\newcommand{\varnow}{\mathsf{var}\mbox{-}\mathsf{now}}
+\newcommand{\laters}{\mathsf{laters}}
+\newcommand{\Now}{\mathsf{Now}}
+\newcommand{\Later}{\mathsf{Later}}
+\newcommand{\zero}{\mathsf{zero}}
+\newcommand{\suc}[1]{\mathsf{suc}\app#1}
+\newcommand{\AgdaMissingDefinition}[1]{#1}
 
 \usepackage{newunicodechar}
-\newunicodechar{λ}{\ensuremath{\mathnormal\lambda}}
 \newunicodechar{←}{\ensuremath{\mathnormal\from}}
@@
 \newunicodechar{↓}{\ensuremath{\mathnormal\downarrow}}
+\newunicodechar{⇓}{\ensuremath{\mathnormal\Downarrow}}
 \newunicodechar{∀}{\ensuremath{\mathnormal\forall}}
 \newunicodechar{∃}{\ensuremath{\exists}}
+\newunicodechar{Π}{\ensuremath{\Pi}}
+\newunicodechar{⊢}{\ensuremath{\vdash}}
+\newunicodechar{⊣}{\ensuremath{\dashv}}
+\newunicodechar{⊧}{\ensuremath{\models}}
+\newunicodechar{⊨}{\ensuremath{\models}}
+\newunicodechar{↦}{\ensuremath{\mapsto}}
 \newunicodechar{⊥}{\ensuremath{\mathnormal\bot}}
@@
 \newunicodechar{ℕ}{\ensuremath{\mathbb N}}
+\newunicodechar{ℤ}{\ensuremath{\mathbb Z}}
+\newunicodechar{𝒫}{\ensuremath{\mathcal P}}
 \newunicodechar{≤}{\ensuremath{\leq}}
-\newunicodechar{∸}{\ensuremath{\div}}
+\newunicodechar{×}{\ensuremath{\times}}
+\newunicodechar{∸}{\ensuremath{\minus}}
 \newunicodechar{′}{\ensuremath{'}}
-\newunicodechar{⇒}{\ensuremath{\Rightarrow}}
 \newunicodechar{₁}{\ensuremath{_1}}
 \newunicodechar{₂}{\ensuremath{_2}}
-\newunicodechar{Σ}{\ensuremath{\Sigma}}
+\newunicodechar{₃}{\ensuremath{_3}}
+\newunicodechar{ⱼ}{\ensuremath{_j}}
 \newunicodechar{⊎}{\ensuremath{\cupdot}}
+\newunicodechar{∪}{\ensuremath{\cup}}
 \newunicodechar{ᵖ}{\ensuremath{^p}}
@@
 \newunicodechar{▷}{\ensuremath{\rhd}}
+\newunicodechar{◁}{\ensuremath{\lhd}}
 \newunicodechar{∈}{\ensuremath{\in}}
+\newunicodechar{⇒}{\ensuremath{\Rightarrow}}
 \newunicodechar{⇔}{\ensuremath{\iff}}
 \newunicodechar{⩦}{\ensuremath{\equiv}} % !!
-\newunicodechar{μ}{\ensuremath{\mu}}
 \newunicodechar{∋}{\ensuremath{\ni}}
@@
 \newunicodechar{∅}{\ensuremath{\emptyset}}
+\newunicodechar{⟨}{\ensuremath{\langle}}
+\newunicodechar{⟩}{\ensuremath{\rangle}}
+\newunicodechar{α}{\ensuremath{\alpha}}
+\newunicodechar{β}{\ensuremath{\beta}}
 \newunicodechar{δ}{\ensuremath{\delta}}
+\newunicodechar{Δ}{\ensuremath{\Delta}}
+\newunicodechar{λ}{\ensuremath{\mathnormal\lambda}}
+\newunicodechar{Λ}{\ensuremath{\mathnormal\Lambda}}
+\newunicodechar{μ}{\ensuremath{\mu}}
+\newunicodechar{ϕ}{\ensuremath{\phi}}
+\newunicodechar{ψ}{\ensuremath{\psi}}
+\newunicodechar{þ}{\ensuremath{\th}}
+\newunicodechar{σ}{\ensuremath{\sigma}}
+\newunicodechar{Σ}{\ensuremath{\Sigma}}
+\newunicodechar{τ}{\ensuremath{\tau}}
+\newunicodechar{♯}{\ensuremath{\sharp}}
+\newunicodechar{♭}{\ensuremath{\flat}}
+\newunicodechar{ℓ}{\ensuremath{\ell}}
+\newunicodechar{ₒ}{\ensuremath{_o}}
+\newunicodechar{ᵃ}{\ensuremath{^a}}
+\newunicodechar{ⁱ}{\ensuremath{^i}}
+\newunicodechar{∎}{\ensuremath{\qed}}
 
@@
 
-%% \author{Lars Th{\o}rv{\"a}ld}
-%% \affiliation{%
-%%   \institution{The Th{\o}rv{\"a}ld Group}
+\author{Philip Wadler}
+\affiliation{%
+  \institution{University of Edinburgh}
   %% \streetaddress{1 Th{\o}rv{\"a}ld Circle}
-%%   \city{Hekla}
-%%   \country{Iceland}}
+  \city{Edinburgh}
+  \country{Scotland}}
 %% \email{larst@affiliation.org}
+
+\author{Peter Thiemann}
+\affiliation{%
+  \institution{University of Freiburg}
+  \city{Freiburg}
+  \country{Germany}
+}
 
@@
 
+\clearpage
+
+\tableofcontents
+
 \section{Introduction}
@@
 Logical relations are a useful technique for proving properties of
-programming languages, from type safety to relational parametricity,
+programming languages, from type safety to relational parametricity
 and noninterference.  Initially, logical relations were applied to
@@
 \citet{Dreyer:2011wl} designed the LSLR logic to streamline the
-reasoning about step indexes. This research program eventually
+reasoning about step indexes. Their research program eventually
 produced Iris~\citep{JUNG:2018aa}, a system that supports a
@@
 assistant~\citep{The-Coq-Development-Team:2004kf,Huet:2016aa}.
+
+TODO: discuss guarded cubical agda
 

### 2023-07-06 papers/cpp2024/paper.tex
check

@@
 \section{Introduction}
+\label{sec:intro}
 
@@
 experience confirm the findings in this paper.
+
+Our goal is to define an operator for recursive predicates and relations
+with syntax that is something like $μᵒ x. R$, where $x$ is the name of the
+recursive relation and $R$ is the definition of the relation, which
+can refer to $x$. We shall prove a fixpoint theorem which states that
+the recursive predicate is equal to its unfolding, something like the
+following.
+\[
+  (μᵒ x. R) \app δ \app a ≡ᵒ R \app δ(x:= μᵒ x. R) \app a
+\]
+where $δ$ is an environment mapping variables to predicates.
+
+Our proof of this fixpoint theorem is inspired by the fixpoint theorem
+of \citet{Appel:2001aa}. In that work, \citet{Appel:2001aa} use
+step-indexing to give a semantic definition of recursive types. Their
+fixpoint theorem proves that a recursive type is equal to its unfolding.
+They define a (semantic) type $\tau$ to be a relation between step indexes and
+syntactic values. They do not define a syntax for types, but instead
+define operators for constructing semantic types as follows.
+\begin{align*}
+  ⊥ &= \{ \} \\
+  ⊤ &= \{ ⟨k,v⟩ \mid k ∈ ℕ\} \\
+  \mathbf{int} &= \{⟨k,n⟩ \mid k ∈ ℕ, n ∈ ℤ \}\\
+  τ₁ × τ₂ &= \{ ⟨k,(v₁,v₂)⟩ \mid ∀j<k. ⟨j,v₁⟩∈τ₁, ⟨j,v₂⟩∈τ₂ \} \\
+  τ₁ → τ₂ &= \{ ⟨k,λx.e⟩ \mid ∀j<k.∀v. ⟨j,v⟩∈τ₁ ⇒ e[v/x] :ⱼ τ₂ \} \\
+  μ F &= \{ ⟨k,v⟩ \mid ⟨k,v⟩ ∈ F^{k\plus 1}(⊥) \}
+    & \text{if } F : \tau \to \tau'
+\end{align*}
+Their fixpoint theorem says that for any well founded $F$,
+\[
+  μ F = F(μF)
+\]
+A well founded function $F$ on types is
+one in which each pair in the output $⟨k,v⟩$ only depends
+on later pairs in the input, that is, pairs of the form $⟨j,v′⟩$
+where $j < k$. \citet{Appel:2001aa} characterize this dependency
+with the help of the $k$-approximation function:
+\[
+  \kapprox(k,τ) = \{ ⟨j,v⟩ \mid j < k, ⟨j,v⟩ ∈ τ\} 
+\]
+They define a \emph{well founded} function $F$ to be one that
+satisfies the following equation.
+\[
+  \kapprox(k \plus 1, F(τ)) = \kapprox(k \plus 1, F(\kapprox(k,τ)))
+\]
+
+Functions over semantic types are not always well founded.  For
+example, the identity function $λα.α$ is not well founded, so one
+cannot apply the fixpoint theorem to the recursive type $μ(λα.α)$
+(which corresponds to the syntactic type $μα.α$).
+On the other hand, the function
+$λα.α×α$ is well founded because of the strict less-than in the
+definition of the $×$ operator. So the fixpont theorem applies to
+$μ(λα.α×α)$.  In general, a function built from the type operators is
+well founded so long as the recursive $α$ only appears underneath a
+type constructor such as $×$ or $→$. \citet{Appel:2001aa} prove this
+fact, which relies on the auxilliary notion of a nonexpansive
+function. In such a function, the output can depend on pairs at the
+current step index as well as later ones. So a \emph{nonexpansive}
+function satisfies the following equation.
+\[
+  \kapprox(k, F(τ)) = \kapprox(k, F(\kapprox(k,τ)))
+\]
+For example, $λα.α$ is nonexpansive and so is $λα.\mathbf{int}$.
+\citet{Appel:2001aa} then prove lemmas about the type constructors.
+For example, regarding products, they prove that if $F$ and $G$
+are nonexpansive functions, then so is $λ α. (F α) × (G α)$.
+
+It is worth noting that \citet{Appel:2001aa} neglect to prove such 
+lemmas for the $μ$ operator itself. For example, given $F : τ₁ → τ₂ → τ₃$
+that is nonexpansive in its first parameter and well founded in
+its second, then $λ α. μ (F α)$ is nonexpansive.
+On the other hand, if $F$ is well founded in both parameters,
+then $λ α. μ (F α)$ is well founded. We shall return to this point later.
+
+Comparing the type operators of \citet{Appel:2001aa} to the logic
+operators of SIL, there are striking similarities. The function type
+operator is quite similar to implication, although one difference is
+that the function type operator uses strict less-than whereas
+implication uses non-strict less-than. The logic introduces the
+``later`` operator, whereas the type operators essentially bake the
+later operator into the type operators through their use of strict
+less-than.
+
+Our definition of recursive predicates will be similar to the
+recursive type of \citet{Appel:2001aa} in that we shall define the
+meaning of a recursive predicate by iteration.  On the other hand, we
+do not want a fixpoint theorem that requires the user of the logic to
+provide a proof that a particular proposition is well
+founded. Instead, we shall introduce a type system for propositions
+that ensure that $μ$ is only applied to well founded propositions, and
+that the proof of well foundedness is provided by our logic operators,
+not by the user of the logic.
 

### 2023-07-19 papers/cpp2024/paper.tex
check
starting on semantic safety of STLC with rec
finished draft of proof theory section

@@
 \usepackage{semantic}
+\usepackage{stmaryrd}
 \usepackage{agda}
@@
 \newcommand{\suc}[1]{\mathsf{suc}\app#1}
+\newcommand{\eff}[1]{\Lbag#1\Rbag}
 \newcommand{\AgdaMissingDefinition}[1]{#1}
@@
 \newunicodechar{↓}{\ensuremath{\mathnormal\downarrow}}
+\newunicodechar{⇑}{\ensuremath{\mathnormal\Uparrow}}
 \newunicodechar{⇓}{\ensuremath{\mathnormal\Downarrow}}
+\newunicodechar{⇒}{\ensuremath{\Rightarrow}}
+\newunicodechar{⇔}{\ensuremath{\iff}}
+\newunicodechar{↠}{\ensuremath{\rightarrow^{\ast}}}
 \newunicodechar{∀}{\ensuremath{\mathnormal\forall}}
@@
 \newunicodechar{∷}{\ensuremath{\mathnormal::}}
+\newunicodechar{⦂}{\ensuremath{:}}  % !!
 \newunicodechar{ℕ}{\ensuremath{\mathbb N}}
@@
 \newunicodechar{₃}{\ensuremath{_3}}
+\newunicodechar{₄}{\ensuremath{_4}}
+\newunicodechar{₅}{\ensuremath{_5}}
 \newunicodechar{ⱼ}{\ensuremath{_j}}
@@
 \newunicodechar{∈}{\ensuremath{\in}}
-\newunicodechar{⇒}{\ensuremath{\Rightarrow}}
-\newunicodechar{⇔}{\ensuremath{\iff}}
 \newunicodechar{⩦}{\ensuremath{\equiv}} % !!
@@
 \newunicodechar{⟩}{\ensuremath{\rangle}}
+\newunicodechar{⟪}{\ensuremath{\langle}}
+\newunicodechar{⟫}{\ensuremath{\langle}}
+\newunicodechar{⟅}{\ensuremath{\Lbag}}
+\newunicodechar{⟆}{\ensuremath{\Rbag}}
+
 \newunicodechar{α}{\ensuremath{\alpha}}
 \newunicodechar{β}{\ensuremath{\beta}}
+\newunicodechar{γ}{\ensuremath{\gamma}}
 \newunicodechar{δ}{\ensuremath{\delta}}
 \newunicodechar{Δ}{\ensuremath{\Delta}}
+\newunicodechar{ξ}{\ensuremath{\xi}}
 \newunicodechar{λ}{\ensuremath{\mathnormal\lambda}}
+\newunicodechar{ƛ}{\ensuremath{\mathnormal\lambda}}
 \newunicodechar{Λ}{\ensuremath{\mathnormal\Lambda}}
 \newunicodechar{μ}{\ensuremath{\mu}}
+\newunicodechar{ν}{\ensuremath{\nu}}
 \newunicodechar{ϕ}{\ensuremath{\phi}}
 \newunicodechar{ψ}{\ensuremath{\psi}}
-\newunicodechar{þ}{\ensuremath{\th}}
+\newunicodechar{þ}{\ensuremath{\chi}} % !!
 \newunicodechar{σ}{\ensuremath{\sigma}}
@@
 \newunicodechar{ⁱ}{\ensuremath{^i}}
+\newunicodechar{ⱽ}{\ensuremath{^V}}
 \newunicodechar{∎}{\ensuremath{\qed}}
+\newunicodechar{■}{\ensuremath{\qed}} % !!
+\newunicodechar{□}{\ensuremath{\qed}} % !!
+\newunicodechar{⦅}{\ensuremath{\llparenthesis}}
+\newunicodechar{⦆}{\ensuremath{\rrparenthesis}}
+\newunicodechar{⟦}{\ensuremath{\llbracket}}
+\newunicodechar{⟧}{\ensuremath{\rrbracket}}
+\newunicodechar{⦉}{\ensuremath{\llparenthesis}} % Z notation...
+\newunicodechar{⦊}{\ensuremath{\rrparenthesis}} % Z notation...
+\newunicodechar{ℰ}{\ensuremath{\mathcal{E}}}
+\newunicodechar{𝒱}{\ensuremath{\mathcal{V}}}
+\newunicodechar{𝓖}{\ensuremath{\mathcal{G}}}
+\newunicodechar{𝓔}{\ensuremath{\mathcal{E}}}
 
@@
   in the Coq proof assistant. This paper describes a shallow embedding
-  of a step-indexed logic in Agda and its application to a semantic
-  type safety proof for an internal language of the Gradually-Typed
-  Lambda Calculus.
+  of a step-indexed modal logic in Agda and its application to a semantic
+  type safety proof for the Simply-Typed Lambda Calculus extended with
+  recursive functions.
 \end{abstract}
@@
 
+\input{latex/cpp2024/STLC.tex}
+\input{latex/cpp2024/STLCBind.tex}
+\input{latex/cpp2024/STLCSafe.tex}
+
 \bibliographystyle{ACM-Reference-Format}

### 2023-07-21 papers/cpp2024/paper.tex
progress on STLC
compatibility lemmas
finished sem type safety
check

@@
 \newcommand{\proj}[2]{#1 \,?\, #2}
-\newcommand{\kapprox}{\mathrm{approx}}
+% \newcommand{\kapprox}[2]{\mathrm{approx}(#1, #2)}
+\newcommand{\kapprox}[2]{↓_{#1}\, #2}
 \newcommand{\varnow}{\mathsf{var}\mbox{-}\mathsf{now}}
@@
 \newunicodechar{∘}{\ensuremath{\circ}}
+\newunicodechar{·}{\ensuremath{\cdot}}
 \newunicodechar{ᵒ}{\ensuremath{^\circ}}
@@
 \newunicodechar{∈}{\ensuremath{\in}}
-\newunicodechar{⩦}{\ensuremath{\equiv}} % !!
+\newunicodechar{⩦}{\ensuremath{\doteq}} % !!
 \newunicodechar{∋}{\ensuremath{\ni}}
@@
 \newunicodechar{⟩}{\ensuremath{\rangle}}
-\newunicodechar{⟪}{\ensuremath{\langle}}
-\newunicodechar{⟫}{\ensuremath{\langle}}
+\newunicodechar{⟪}{\ensuremath{\langle\!\!\langle}}
+\newunicodechar{⟫}{\ensuremath{\rangle\!\!\rangle}}
 \newunicodechar{⟅}{\ensuremath{\Lbag}}
@@
 \newunicodechar{λ}{\ensuremath{\mathnormal\lambda}}
-\newunicodechar{ƛ}{\ensuremath{\mathnormal\lambda}}
+\newunicodechar{ƛ}{\ensuremath{\mathnormal\lambdabar}}
 \newunicodechar{Λ}{\ensuremath{\mathnormal\Lambda}}
 \newunicodechar{μ}{\ensuremath{\mu}}
+\newunicodechar{π}{\ensuremath{\pi}}
 \newunicodechar{ν}{\ensuremath{\nu}}
@@
 \newunicodechar{ⱽ}{\ensuremath{^V}}
-\newunicodechar{∎}{\ensuremath{\qed}}
-\newunicodechar{■}{\ensuremath{\qed}} % !!
-\newunicodechar{□}{\ensuremath{\qed}} % !!
+\newunicodechar{∎}{\ensuremath{\blacksquare}} % !!
+\newunicodechar{■}{\ensuremath{\blacksquare}}
+\newunicodechar{□}{\ensuremath{\Box}}
+\newunicodechar{•}{\ensuremath{\bullet}}
 \newunicodechar{⦅}{\ensuremath{\llparenthesis}}
@@
   languages~\citep{Appel:2001aa,Ahmed:2006aa}, and their application
-  can be streamlined by use of modal logics such as
-  LSLR~\citep{Dreyer:2011wl}.  The Iris system~\citep{JUNG:2018aa}
-  provides support for reasoning about step-indexed logical relations
-  in the Coq proof assistant. This paper describes a shallow embedding
-  of a step-indexed modal logic in Agda and its application to a semantic
+  can be streamlined by use of step-indexed modal logics such as
+  LSLR~\citep{Dreyer:2011wl}.  This paper describes an embedding of a
+  step-indexed modal logic in Agda and its application to a semantic
   type safety proof for the Simply-Typed Lambda Calculus extended with
@@
 Logical relations are a useful technique for proving properties of
-programming languages, from type safety to relational parametricity
-and noninterference.  Initially, logical relations were applied to
-terminating languages such as the Simply Typed Lambda
-Calculus~\citep{Tait:1967aa} and System F~\citep{GIRARD72}, but
-\citet{Appel:2001aa} extended the technique to handle languages with
-nontermination by introducing a step index into the logical
-relation. After some experience with applications of step-indexed
-logical
-relations~\citep{Ahmed:2004eu,Ahmed:2009aa,Neis:2009fk,Hur:2011aa},
-\citet{Dreyer:2011wl} designed the LSLR logic to streamline the
-reasoning about step indexes. Their research program eventually
-produced Iris~\citep{JUNG:2018aa}, a system that supports a
-combination of step indexed logical relations and separation logic
-within the Coq proof
+programming languages, from type safety~\citep{Timany:2022aa} to
+relational parametricity~\citep{REYNOLDS83}, and
+noninterference~\citep{heintze1998slam}.  Initially, logical relations
+were applied to terminating languages such as the Simply Typed Lambda
+Calculus~\citep{Tait:1967aa} and System F~\citep{GIRARD72}.
+\citet{Appel:2001aa} introduced step-indexed logical relations to
+handle languages with recursive types.  \citet{Dreyer:2011wl} designed
+the LSLR logic to streamline the reasoning about step indexed logical
+relations.  This work eventually led to Iris~\citep{JUNG:2018aa}, a
+system that supports a combination of step indexing and separation
+logic within the Coq proof
 assistant~\citep{The-Coq-Development-Team:2004kf,Huet:2016aa}.
 
-TODO: discuss guarded cubical agda
-
-The author is interested in step-indexed logical relations for
+The authors are interested in using step-indexed logical relations for
 purposes of developing mechanized proofs in Agda of the Gradual
@@
 for gradually-typed polymorphic
-languages~\citep{Ahmed:2011fk,Ahmed:2017aa,Igarashi:2017aa,New:2019ab,Labrada:2020tk}.
-To streamline such developments in Agda, we need support for a
-step-indexed logic in Agda. This paper fills that need by showing how
-to build a shallow embedding of a step-indexed logic in Agda, a system
-that we abbreviate as SIL. To evaluate the utility of SIL, we present
-a case study of proving semantic type safety of a simple cast
-calculus, i.e., an internal language for the Gradually-Typed Lambda
-Calculus (GTLC)~\citep{Siek:2006bh,Siek:2007qy}. We find that SIL
-streamlines the definition of logical relations and is helpful in many
-of the proofs, but there is room for improvement regarding the
-ergonomics of interactive proof development using SIL.  Limitations in
-Agda's inference algorithm sometimes leads to verbosity in SIL
-proofs. It is worth noting that the authors have also proved the
-Gradual Guarantee for the GTLC using SIL and the lesssons from that
-experience confirm the findings in this paper.
+languages~\citep{Ahmed:2011fk,Ahmed:2017aa,Igarashi:2017aa,New:2019ab,Labrada:2020tk}. One
+option would be to use Guarded Cubical Agda~\citep{Veltri:2020aa}, but
+both Cubical Agda~\citep{Vezzosi:2019aa} and its extension with
+guarded fixpoints are experimental features of Agda that may not be
+supported indefinitely.  We would prefer to have a step-indexed logic
+that is embedded in (vanilla) Agda.
 
-Our goal is to define an operator for recursive predicates and relations
-with syntax that is something like $μᵒ x. R$, where $x$ is the name of the
-recursive relation and $R$ is the definition of the relation, which
-can refer to $x$. We shall prove a fixpoint theorem which states that
-the recursive predicate is equal to its unfolding, something like the
-following.
+This paper fills that need by showing how to build a step-indexed
+logic in Agda, a system that we abbreviate as SIL.  This paper is a
+literate Agda file; all of the definitions and proofs have been
+checked with Agda 2.6.2.2. 
+%
+Most of the challenge in defining a step-indexed logic is in the
+definition of recursive predicates and their fixpoint theorem. The
+step-indexed models of \citet{Appel:2001aa} rely on an approximation
+operator, here written as $↓_n$, to define the semantic criteria for
+valid recursive definitions.  For example, they define a nonexpansive
+function to be one that satisfies
 \[
-  (μᵒ x. R) \app δ \app a ≡ᵒ R \app δ(x:= μᵒ x. R) \app a
+  ↓_n f(x) = ↓_n f(↓_n x)
 \]
-where $δ$ is an environment mapping variables to predicates.
+The step-indexed models of \citet{JUNG:2018aa} instead rely on
+a step-indexed notion of equality. For example, they define a
+nonexpansive function as one that satisfies
+\[
+  x \stackrel{n}{=} y \text{ implies } f(x) \stackrel{n}{=} f(y)
+\]
+We are not aware of any inherent reason to prefer one approach to the
+other, so we choose the approach of \citet{Appel:2001aa} because they
+provide perspicuous proofs of their fixpoint theorem.  So our approach
+to SIL can be seen as an implementation of the LSLR logic of
+\citet{Dreyer:2011wl} using the foundations of
+\citet{Appel:2001aa}. In particular, SIL defines the recursive
+predicate of SIL using function iteration, like this:
+\[
+  ⟦ μ F ⟧ \, x \, n = F^{n\plus 1}\, \top \, x \, n
+\]
+whereas LSLR uses substitution in the semantics of recursive types:
+\[
+   ⟦ μ r. R ⟧ \, x \, n = ⟦ R[μr.R/r] ⟧\,x\,n
+\]
 
-Our proof of this fixpoint theorem is inspired by the fixpoint theorem
-of \citet{Appel:2001aa}. In that work, \citet{Appel:2001aa} use
-step-indexing to give a semantic definition of recursive types. Their
-fixpoint theorem proves that a recursive type is equal to its unfolding.
-They define a (semantic) type $\tau$ to be a relation between step indexes and
-syntactic values. They do not define a syntax for types, but instead
-define operators for constructing semantic types as follows.
+However, adapting the work of \citet{Appel:2001aa} to our setting was
+nontrivial because (1) their model is for recursive types, not a modal
+logic, and (2) their proofs were incomplete (which we learned part way
+through our development of SIL). To enable the nesting of recursive
+types inside other recursive types, one must prove that the $μ$ type
+constructor is nonexpansive and well founded, but \citet{Appel:2001aa}
+does not prove this (they prove those properties for the other type
+constructors).  We discuss this problem and its remedy in sections
+\ref{sec:appel-mcallester} and \ref{sec:open-propositions}.
+
+We organize this paper as follows. We review Appel an McAllester's
+work in Section~\ref{sec:appel-mcallester}, then introduce our
+representation of step-indexed propositions and predicates in Agda in
+Sections \ref{sec:propositions} and \ref{sec:predicates}.  We
+formalize several important concepts in
+Section~\ref{sec:fun-approx-iter}: functionals, approximation, and
+iteration, and prove some basic results about them.  We construct all
+of the logical connectives of SIL in
+Section~\ref{sec:open-propositions}, including the definition of
+recursive predicates and their fixpoint theorem.  The proof theory of
+SIL is developed in Section~\ref{sec:proofs}, where we construct the
+proof rules of this modal logic.  To evaluate the utility of SIL,
+Section~\ref{sec:STLC} presents a case study of proving semantic type
+safety for the Simply Typed Lambda Calculus extended with recursive
+functions. We find that SIL streamlines the definition of logical
+relations and is helpful in many of the proofs, but there is room for
+improvement regarding the ergonomics of interactive proof development
+using SIL.  Limitations in Agda's inference algorithm sometimes leads
+to verbosity in SIL proofs.
+
+%% Our goal is to define an operator for recursive predicates and relations
+%% with syntax that is something like $μᵒ x. R$, where $x$ is the name of the
+%% recursive relation and $R$ is the definition of the relation, which
+%% can refer to $x$. We shall prove a fixpoint theorem which states that
+%% the recursive predicate is equal to its unfolding, something like the
+%% following.
+%% \[
+%%   (μᵒ x. R) \app δ \app a ≡ᵒ R \app δ(x:= μᵒ x. R) \app a
+%% \]
+%% where $δ$ is an environment mapping variables to predicates.
+
+\section{Review of Appel and McAllester's Fixpoint Theorem}
+\label{sec:appel-mcallester}
+
+Our proof of the fixpoint theorem for SIL is inspired by the fixpoint
+theorem of \citet{Appel:2001aa}. Their fixpoint theorem proves that a
+recursive type is equal to its unfolding.  They define a (semantic)
+type $\tau$ to be a relation between step indexes and syntactic
+values. They do not define a syntax for types, but instead define
+operators for constructing semantic types as follows.
 \begin{align*}
@@
 \[
-  \kapprox(k,τ) = \{ ⟨j,v⟩ \mid j < k, ⟨j,v⟩ ∈ τ\} 
+  \kapprox{k}{τ} = \{ ⟨j,v⟩ \mid j < k, ⟨j,v⟩ ∈ τ\} 
 \]
@@
 \[
-  \kapprox(k \plus 1, F(τ)) = \kapprox(k \plus 1, F(\kapprox(k,τ)))
+  \kapprox{k \plus 1}{F(τ)} = \kapprox{k \plus 1}{F(\kapprox{k}{τ})}
 \]
@@
 \[
-  \kapprox(k, F(τ)) = \kapprox(k, F(\kapprox(k,τ)))
+  \kapprox{k}{F(τ)} = \kapprox{k}{F(\kapprox{k}{τ})}
 \]
@@
 
-It is worth noting that \citet{Appel:2001aa} neglect to prove such 
-lemmas for the $μ$ operator itself. For example, given $F : τ₁ → τ₂ → τ₃$
-that is nonexpansive in its first parameter and well founded in
-its second, then $λ α. μ (F α)$ is nonexpansive.
-On the other hand, if $F$ is well founded in both parameters,
-then $λ α. μ (F α)$ is well founded. We shall return to this point later.
+However, \citet{Appel:2001aa} neglect to prove such lemmas for the $μ$
+operator itself. For example, given $F : τ₁ → τ₂ → τ₃$ that is
+nonexpansive in its first parameter and well founded in its second,
+then $λ α. μ (F α)$ is nonexpansive.  On the other hand, if $F$ is
+well founded in both parameters, then $λ α. μ (F α)$ is well founded.
 
@@
 that the function type operator uses strict less-than whereas
-implication uses non-strict less-than. The logic introduces the
-``later`` operator, whereas the type operators essentially bake the
-later operator into the type operators through their use of strict
+implication uses non-strict less-than. The logic introduces the later
+operator, whereas the type operators essentially bake the later
+operator into the type operators through their use of strict
 less-than.
 
-Our definition of recursive predicates will be similar to the
-recursive type of \citet{Appel:2001aa} in that we shall define the
-meaning of a recursive predicate by iteration.  On the other hand, we
-do not want a fixpoint theorem that requires the user of the logic to
-provide a proof that a particular proposition is well
-founded. Instead, we shall introduce a type system for propositions
-that ensure that $μ$ is only applied to well founded propositions, and
-that the proof of well foundedness is provided by our logic operators,
-not by the user of the logic.
-
-TODO: ROAD MAP
-
-%\section{A Step-Indexed Logic (SIL)}
-
-%% Figure~\ref{fig:SIL}
-
-%% \begin{figure}
-%%   \raggedright
-%% Syntax
-%%   \begin{syntax}
-%%     \text{Propositions} & P,Q &::= & \top \mid \bot \mid P \land Q
-%%     \mid P \lor Q \mid P \Rightarrow Q \mid \forall X.P \mid \exists X.P
-%%     \mid \overline{M} \in R\\
-%%     \text{Relations} & R,S & ::= & r \mid \mu r. R \mid \{ \overline{x} \mid P \}
-%%   \end{syntax}
-%% Meaning of Propositions
-%% \begin{align*}
-%%   \sem{ P } \rho\; 0 &= \top \\
-%%   \sem{ \top } \rho(suc \app n) &= \top \\
-%%   \sem{ \bot } \rho (suc \app n) &= \bot \\
-%%   \sem{ P \land Q } \rho (suc \app n) &=
-%%     \sem{ P } \rho n \land \sem{ Q } \rho n \\
-%%   \sem{ P \lor Q } \rho (suc \app n) &=
-%%     \sem{ P } \rho n \lor \sem{ Q } \rho n \\
-%%   \sem{ P \Rightarrow Q } \rho (suc \app n) &=
-%%     \forall k \leq suc \app n,
-%%     \sem{ P } \rho k \implies \sem{ Q } \rho k \\
-%%   \sem{ \forall X.P} \rho \app (suc \app n) &= \forall a, \sem{P[a/X]} \rho (suc \app n)\\
-%%   \sem{ \exists X.P} \rho \app (suc \app n) &= \exists a, \sem{P[a/X]} \rho (suc \app n)\\
-%%   \sem{ \overline{M} \in R } \rho (suc \app n) &=
-%%     \sem{ R } \rho (suc \app n) \app \overline{M}\\
-%%   \sem{ \rhd P } \rho (suc \app n) &=
-%%     \sem{ P } \rho n 
-%% \end{align*}
-%% Meaning of Relations
-%% \begin{align*}
-%%   \sem{ R } \rho\, 0 \app \overline{M} &= \top \\
-%%   \sem{ r } \rho \app (suc \app n) \app \overline{M} &=
-%%     \rho(r) (suc \app n) \app \overline{M} \\
-%%   \sem{ \mu r. R } \rho \app (suc \app n)\app \overline{M} &=
-%%   F^{suc \app n}(\top) (suc \app n) \app \overline{M} \\
-%%   & \text{ where } F(X) = \sem{ R } \rho(r := X) \\
-%%   \sem{ \{ \overline{x} \mid P \} } \rho (suc \app n) \app \overline{M} &= 
-%%     \sem{ P[\overline{x} := \overline{M}] } \rho (suc \app n)
-%% \end{align*}
-%%   \caption{Step-Indexed Logic (SIL)}
-%%   \label{fig:SIL}
-%% \end{figure}
-
-%\section{Case Study: Semantic Type Safety of a Cast Calculus}
-
-%% Figure~\ref{fig:CC}
-
-%% \begin{figure}
-%%   \raggedright Syntax
-%%   \begin{syntax}
-%%     \text{Types} & A,B,C,D &::=& \Bool \mid A \to B \mid \dyn \\
-%%     \text{Ground Types} & G,H & ::= & \mathtt{bool} \mid \dyn \to \dyn \\
-%%     \text{Booleans} & b & \in & \mathbb{B} \\
-%%     \text{Terms} & L,M,N & ::= & x \mid b \mid \lam{x}N \mid L \app M
-%%       \mid \inj{M}{G} \mid \proj{M}{H} \mid \mathtt{error} \\
-%%       \text{Values} & V,W & ::= & \lam{x}N \mid \inj{V}{G}
-%%   \end{syntax}
-%%   Type System
-%%   \begin{gather*}
-%%     \inference{}{\Gamma \vdash x : \Gamma(x)}
-%%     \quad
-%%     \inference{}{\Gamma \vdash b : \Bool}
-%%     \quad
-%%     \inference{\Gamma, x : A \vdash N : B}
-%%               {\Gamma \vdash \lam{x} N : A \to B}
-%%     \quad              
-%%     \inference{\Gamma \vdash L : A \to B & \Gamma \vdash M : A}
-%%           {\Gamma \vdash L \app M : B} \\[2ex]
-%%    \inference{\Gamma \vdash M : G}
-%%           {\Gamma \vdash \inj{M}{G} : \dyn}
-%%    \quad
-%%    \inference{\Gamma \vdash M : \dyn}
-%%              {\Gamma \vdash \proj{M}{H} : H}
-%%    \quad
-%%    \inference{}
-%%              {\Gamma \vdash \mathtt{error} : A}
-%%   \end{gather*}
-%% Reduction
-%% \begin{align*}
-%%   (\lam{x} N)\app W & \reduce N[x:=W] \\
-%%   \proj{\inj{V}{G}}{G} & \reduce V \\
-%%   \proj{\inj{V}{G}}{H} & \reduce \texttt{error}
-%%        \qquad\qquad\qquad\text{if } G \neq H 
-%% \end{align*}
-
-%%   \caption{Internal language (cast calculus) for the Gradually-Typed
-%%     Lambda Calculus.}
-%%   \label{fig:CC}
-%% \end{figure}
-
-%% \begin{align*}
-%%     \semE{\_} &: \mathsf{Type} \to \mathsf{Term} \to \mathsf{Set} \\
-%%     \semE{A} M &= (\semV{A} M \lor \mathsf{reducible}\ M \lor M=\mathtt{blame})
-%%                 \land (\forall N, (M \reduce N) \Rightarrow \rhd \semE{A} N)
-%% \end{align*}
+As noted above, our semantics of recursive predicates is similar to
+the recursive type of \citet{Appel:2001aa} in that we define recursive
+predicates via iteration.
+%
+However, we do not want to require users of the logic to have to prove
+the well foundedness of their recursive formulas.  Instead, we
+introduce a type system for propositions that ensure that $μ$ is only
+applied to well founded formulas, and that the proof of well
+foundedness is provided by our logical connectives, not by the user of
+the logic.
 
@@
 
+\section*{Appendix}
+
+\input{latex/cpp2024/STLCDeterministic.tex}
+
 \end{document}

### 2023-07-24 papers/July2024/paper.tex
rough draft

--- 2023-07-21 papers/cpp2024/paper.tex
@@
   can be streamlined by use of step-indexed modal logics such as
-  LSLR~\citep{Dreyer:2011wl}.  This paper describes an embedding of a
-  step-indexed modal logic in Agda and its application to a semantic
-  type safety proof for the Simply-Typed Lambda Calculus extended with
-  recursive functions.
+  LSLR~\citep{Dreyer:2011wl}.  This paper describes a mechanization of
+  a step-indexed modal logic in the Agda proof assistant and its
+  application to a semantic type safety proof for the Simply-Typed
+  Lambda Calculus extended with recursive functions. The step-indexed
+  logic supports guarded recursive predicates via the later modality
+  and provides a fixpoint theorem adapted from \citet{Appel:2001aa}.
 \end{abstract}
@@
 
-\input{latex/cpp2024/StepIndexedLogic.tex}
+\input{latex/July2024/StepIndexedLogic.tex}
 
-\input{latex/cpp2024/STLC.tex}
-\input{latex/cpp2024/STLCBind.tex}
-\input{latex/cpp2024/STLCSafe.tex}
+\input{latex/July2024/STLC.tex}
+\input{latex/July2024/STLCBind.tex}
+\input{latex/July2024/STLCSafe.tex}
 
@@
 
+\appendix
+\addcontentsline{toc}{section}{Appendix}
 \section*{Appendix}
 
-\input{latex/cpp2024/STLCDeterministic.tex}
+\input{latex/July2024/STLCDeterministic.tex}
 

### 2023-07-26 papers/July2024/paper.tex
check

@@
 
-\section{Review of Appel and McAllester's Fixpoint Theorem}
-\label{sec:appel-mcallester}
-
-Our proof of the fixpoint theorem for SIL is inspired by the fixpoint
-theorem of \citet{Appel:2001aa}. Their fixpoint theorem proves that a
-recursive type is equal to its unfolding.  They define a (semantic)
-type $\tau$ to be a relation between step indexes and syntactic
-values. They do not define a syntax for types, but instead define
-operators for constructing semantic types as follows.
-\begin{align*}
-  ⊥ &= \{ \} \\
-  ⊤ &= \{ ⟨k,v⟩ \mid k ∈ ℕ\} \\
-  \mathbf{int} &= \{⟨k,n⟩ \mid k ∈ ℕ, n ∈ ℤ \}\\
-  τ₁ × τ₂ &= \{ ⟨k,(v₁,v₂)⟩ \mid ∀j<k. ⟨j,v₁⟩∈τ₁, ⟨j,v₂⟩∈τ₂ \} \\
-  τ₁ → τ₂ &= \{ ⟨k,λx.e⟩ \mid ∀j<k.∀v. ⟨j,v⟩∈τ₁ ⇒ e[v/x] :ⱼ τ₂ \} \\
-  μ F &= \{ ⟨k,v⟩ \mid ⟨k,v⟩ ∈ F^{k\plus 1}(⊥) \}
-    & \text{if } F : \tau \to \tau'
-\end{align*}
-Their fixpoint theorem says that for any well founded $F$,
-\[
-  μ F = F(μF)
-\]
-A well founded function $F$ on types is
-one in which each pair in the output $⟨k,v⟩$ only depends
-on later pairs in the input, that is, pairs of the form $⟨j,v′⟩$
-where $j < k$. \citet{Appel:2001aa} characterize this dependency
-with the help of the $k$-approximation function:
-\[
-  \kapprox{k}{τ} = \{ ⟨j,v⟩ \mid j < k, ⟨j,v⟩ ∈ τ\} 
-\]
-They define a \emph{well founded} function $F$ to be one that
-satisfies the following equation.
-\[
-  \kapprox{k \plus 1}{F(τ)} = \kapprox{k \plus 1}{F(\kapprox{k}{τ})}
-\]
-
-Functions over semantic types are not always well founded.  For
-example, the identity function $λα.α$ is not well founded, so one
-cannot apply the fixpoint theorem to the recursive type $μ(λα.α)$
-(which corresponds to the syntactic type $μα.α$).
-On the other hand, the function
-$λα.α×α$ is well founded because of the strict less-than in the
-definition of the $×$ operator. So the fixpont theorem applies to
-$μ(λα.α×α)$.  In general, a function built from the type operators is
-well founded so long as the recursive $α$ only appears underneath a
-type constructor such as $×$ or $→$. \citet{Appel:2001aa} prove this
-fact, which relies on the auxilliary notion of a nonexpansive
-function. In such a function, the output can depend on pairs at the
-current step index as well as later ones. So a \emph{nonexpansive}
-function satisfies the following equation.
-\[
-  \kapprox{k}{F(τ)} = \kapprox{k}{F(\kapprox{k}{τ})}
-\]
-For example, $λα.α$ is nonexpansive and so is $λα.\mathbf{int}$.
-\citet{Appel:2001aa} then prove lemmas about the type constructors.
-For example, regarding products, they prove that if $F$ and $G$
-are nonexpansive functions, then so is $λ α. (F α) × (G α)$.
-
-However, \citet{Appel:2001aa} neglect to prove such lemmas for the $μ$
-operator itself. For example, given $F : τ₁ → τ₂ → τ₃$ that is
-nonexpansive in its first parameter and well founded in its second,
-then $λ α. μ (F α)$ is nonexpansive.  On the other hand, if $F$ is
-well founded in both parameters, then $λ α. μ (F α)$ is well founded.
-
-Comparing the type operators of \citet{Appel:2001aa} to the logic
-operators of SIL, there are striking similarities. The function type
-operator is quite similar to implication, although one difference is
-that the function type operator uses strict less-than whereas
-implication uses non-strict less-than. The logic introduces the later
-operator, whereas the type operators essentially bake the later
-operator into the type operators through their use of strict
-less-than.
-
-As noted above, our semantics of recursive predicates is similar to
-the recursive type of \citet{Appel:2001aa} in that we define recursive
-predicates via iteration.
-%
-However, we do not want to require users of the logic to have to prove
-the well foundedness of their recursive formulas.  Instead, we
-introduce a type system for propositions that ensure that $μ$ is only
-applied to well founded formulas, and that the proof of well
-foundedness is provided by our logical connectives, not by the user of
-the logic.
-
-\input{latex/July2024/StepIndexedLogic.tex}
-
 \input{latex/July2024/STLC.tex}
@@
 \input{latex/July2024/STLCSafe.tex}
+
+\input{review-appel.tex}
+
+\input{latex/July2024/StepIndexedLogic.tex}
 

### 2023-07-26 papers/July2023/paper.tex
check

--- 2023-07-26 papers/July2024/paper.tex
@@
 checked with Agda 2.6.2.2.
-%
+
+%% Our goal is to define an operator for recursive predicates and relations
+%% with syntax that is something like $μᵒ x. R$, where $x$ is the name of the
+%% recursive relation and $R$ is the definition of the relation, which
+%% can refer to $x$. We shall prove a fixpoint theorem which states that
+%% the recursive predicate is equal to its unfolding, something like the
+%% following.
+%% \[
+%%   (μᵒ x. R) \app δ \app a ≡ᵒ R \app δ(x:= μᵒ x. R) \app a
+%% \]
+%% where $δ$ is an environment mapping variables to predicates.
+
+UNDER CONSTRUCTION
+
+To evaluate the utility of SIL, Section~\ref{sec:STLC} presents a case
+study of proving semantic type safety for the Simply Typed Lambda
+Calculus extended with recursive functions. We find that SIL
+streamlines the definition of logical relations and is helpful in many
+of the proofs, but there is room for improvement regarding the
+ergonomics of interactive proof development using SIL.  Limitations in
+Agda's inference algorithm sometimes leads to verbosity in SIL proofs.
+
+\input{latex/July2023/SILIntro.tex}
+
+\input{latex/July2023/STLC.tex}
+\input{latex/July2023/STLCBind.tex}
+\input{latex/July2023/STLCSafe.tex}
+
+\section{Constructing a Step-Indexed Logic}
+
 Most of the challenge in defining a step-indexed logic is in the
@@
 
-We organize this paper as follows. We review Appel an McAllester's
-work in Section~\ref{sec:appel-mcallester}, then introduce our
-representation of step-indexed propositions and predicates in Agda in
-Sections \ref{sec:propositions} and \ref{sec:predicates}.  We
-formalize several important concepts in
+We organize the rest of the article as follows. We review Appel an
+McAllester's work in Section~\ref{sec:appel-mcallester}, then
+introduce our representation of step-indexed propositions and
+predicates in Agda in Sections \ref{sec:propositions} and
+\ref{sec:predicates}.  We formalize several important concepts in
 Section~\ref{sec:fun-approx-iter}: functionals, approximation, and
@@
 SIL is developed in Section~\ref{sec:proofs}, where we construct the
-proof rules of this modal logic.  To evaluate the utility of SIL,
-Section~\ref{sec:STLC} presents a case study of proving semantic type
-safety for the Simply Typed Lambda Calculus extended with recursive
-functions. We find that SIL streamlines the definition of logical
-relations and is helpful in many of the proofs, but there is room for
-improvement regarding the ergonomics of interactive proof development
-using SIL.  Limitations in Agda's inference algorithm sometimes leads
-to verbosity in SIL proofs.
-
-%% Our goal is to define an operator for recursive predicates and relations
-%% with syntax that is something like $μᵒ x. R$, where $x$ is the name of the
-%% recursive relation and $R$ is the definition of the relation, which
-%% can refer to $x$. We shall prove a fixpoint theorem which states that
-%% the recursive predicate is equal to its unfolding, something like the
-%% following.
-%% \[
-%%   (μᵒ x. R) \app δ \app a ≡ᵒ R \app δ(x:= μᵒ x. R) \app a
-%% \]
-%% where $δ$ is an environment mapping variables to predicates.
-
-\input{latex/July2024/STLC.tex}
-\input{latex/July2024/STLCBind.tex}
-\input{latex/July2024/STLCSafe.tex}
+proof rules of this modal logic.
 
@@
 
-\input{latex/July2024/StepIndexedLogic.tex}
+\input{latex/July2023/StepIndexedLogic.tex}
 
@@
 
-\input{latex/July2024/STLCDeterministic.tex}
+\input{latex/July2023/STLCDeterministic.tex}
 

### 2023-07-26 papers/July2023/review-appel.tex
check

--- 2023-07-21 papers/cpp2024/paper.tex
@@
-%%
-%% This is file `sample-acmsmall.tex',
-%% generated with the docstrip utility.
-%%
-%% The original source files were:
-%%
-%% samples.dtx  (with options: `acmsmall')
-%% 
-%% IMPORTANT NOTICE:
-%% 
-%% For the copyright see the source file.
-%% 
-%% Any modified versions of this file must be renamed
-%% with new filenames distinct from sample-acmsmall.tex.
-%% 
-%% For distribution of the original source see the terms
-%% for copying and modification in the file samples.dtx.
-%% 
-%% This generated file may be distributed as long as the
-%% original source files, as listed above, are part of the
-%% same distribution. (The sources need not necessarily be
-%% in the same archive or directory.)
-%%
-%%
-%% Commands for TeXCount
-%TC:macro \cite [option:text,text]
-%TC:macro \citep [option:text,text]
-%TC:macro \citet [option:text,text]
-%TC:envir table 0 1
-%TC:envir table* 0 1
-%TC:envir tabular [ignore] word
-%TC:envir displaymath 0 word
-%TC:envir math 0 word
-%TC:envir comment 0 0
-%%
-%%
-%% The first command in your LaTeX source must be the \documentclass
-%% command.
-%%
-%% For submission and review of your manuscript please change the
-%% command to \documentclass[manuscript, screen, review]{acmart}.
-%%
-%% When submitting camera ready or to TAPS, please change the command
-%% to \documentclass[sigconf]{acmart} or whichever template is required
-%% for your publication.
-%%
-%%
-\documentclass[acmsmall]{acmart}
-\usepackage{semantic}
-\usepackage{stmaryrd}
-\usepackage{agda}
-\usepackage{MnSymbol}
-
-%%
-%% \BibTeX command to typeset BibTeX logo in the docs
-\AtBeginDocument{%
-  \providecommand\BibTeX{{%
-    Bib\TeX}}}
-
-%% Rights management information.  This information is sent to you
-%% when you complete the rights form.  These commands have SAMPLE
-%% values in them; it is your responsibility as an author to replace
-%% the commands and values with those provided to you when you
-%% complete the rights form.
-\setcopyright{acmcopyright}
-\copyrightyear{2023}
-\acmYear{2023}
-\acmDOI{XXXXXXX.XXXXXXX}
-
-%%
-%% These commands are for a JOURNAL article.
-%% \acmJournal{JACM}
-%% \acmVolume{37}
-%% \acmNumber{4}
-%% \acmArticle{111}
-%% \acmMonth{8}
-
-%%
-%% Submission ID.
-%% Use this when submitting an article to a sponsored event. You'll
-%% receive a unique submission ID from the organizers
-%% of the event, and this ID should be used as the parameter to this command.
-%%\acmSubmissionID{123-A56-BU3}
-
-%%
-%% For managing citations, it is recommended to use bibliography
-%% files in BibTeX format.
-%%
-%% You can then either use BibTeX with the ACM-Reference-Format style,
-%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
-%% support for advanced citation of software artefact from the
-%% biblatex-software package, also separately available on CTAN.
-%%
-%% Look at the sample-*-biblatex.tex files for templates showcasing
-%% the biblatex styles.
-%%
-
-%%
-%% The majority of ACM publications use numbered citations and
-%% references.  The command \citestyle{authoryear} switches to the
-%% "author year" style.
-%%
-%% If you are preparing content for an event
-%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
-%% citations and references.
-%% Uncommenting
-%% the next command will enable that style.
-\citestyle{acmauthoryear}
-
-\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
-\newcommand{\semV}[1]{\mathcal{V}\sem{#1}}
-\newcommand{\semE}[1]{\mathcal{E}\sem{#1}}
-\newcommand{\ba}{\begin{array}}
-\newcommand{\ea}{\end{array}}
-\newenvironment{stack}{\ba{@{}l@{}}}{\ea}
-\newenvironment{branch}{\left\{\ba{@{}l@{\qquad}l@{}}}{\ea\right\}}
-\newenvironment{syntax}{\[\ba{l@{\;\;}lcl}}{\ea\]}
-\newcommand{\dotspace}{.\,}
-\newcommand{\key}[1]{\ensuremath{\mathtt{#1}}}
-\newcommand{\dyn}{\star}
-\newcommand{\Dyn}{\ensuremath{\dyn}}
-\newcommand{\Int}{\key{int}}
-\newcommand{\Bool}{\key{bool}}
-\newcommand{\lam}[1]{\lambda #1 \dotspace}
-\newcommand{\app}{\,}
-\newcommand{\reduce}{\longrightarrow}
-\newcommand{\inj}[2]{#1 \,!\, #2}
-\newcommand{\proj}[2]{#1 \,?\, #2}
-% \newcommand{\kapprox}[2]{\mathrm{approx}(#1, #2)}
-\newcommand{\kapprox}[2]{↓_{#1}\, #2}
-\newcommand{\varnow}{\mathsf{var}\mbox{-}\mathsf{now}}
-\newcommand{\laters}{\mathsf{laters}}
-\newcommand{\Now}{\mathsf{Now}}
-\newcommand{\Later}{\mathsf{Later}}
-\newcommand{\zero}{\mathsf{zero}}
-\newcommand{\suc}[1]{\mathsf{suc}\app#1}
-\newcommand{\eff}[1]{\Lbag#1\Rbag}
-\newcommand{\AgdaMissingDefinition}[1]{#1}
-
-\usepackage{newunicodechar}
-\newunicodechar{←}{\ensuremath{\mathnormal\from}}
-\newunicodechar{→}{\ensuremath{\mathnormal\to}}
-\newunicodechar{↓}{\ensuremath{\mathnormal\downarrow}}
-\newunicodechar{⇑}{\ensuremath{\mathnormal\Uparrow}}
-\newunicodechar{⇓}{\ensuremath{\mathnormal\Downarrow}}
-\newunicodechar{⇒}{\ensuremath{\Rightarrow}}
-\newunicodechar{⇔}{\ensuremath{\iff}}
-\newunicodechar{↠}{\ensuremath{\rightarrow^{\ast}}}
-\newunicodechar{∀}{\ensuremath{\mathnormal\forall}}
-\newunicodechar{∃}{\ensuremath{\exists}}
-\newunicodechar{Π}{\ensuremath{\Pi}}
-\newunicodechar{⊢}{\ensuremath{\vdash}}
-\newunicodechar{⊣}{\ensuremath{\dashv}}
-\newunicodechar{⊧}{\ensuremath{\models}}
-\newunicodechar{⊨}{\ensuremath{\models}}
-\newunicodechar{↦}{\ensuremath{\mapsto}}
-\newunicodechar{⊥}{\ensuremath{\mathnormal\bot}}
-\newunicodechar{⊤}{\ensuremath{\mathnormal\top}}
-\newunicodechar{∷}{\ensuremath{\mathnormal::}}
-\newunicodechar{⦂}{\ensuremath{:}}  % !!
-\newunicodechar{ℕ}{\ensuremath{\mathbb N}}
-\newunicodechar{ℤ}{\ensuremath{\mathbb Z}}
-\newunicodechar{𝒫}{\ensuremath{\mathcal P}}
-\newunicodechar{≤}{\ensuremath{\leq}}
-\newunicodechar{×}{\ensuremath{\times}}
-\newunicodechar{∸}{\ensuremath{\minus}}
-\newunicodechar{′}{\ensuremath{'}}
-\newunicodechar{₁}{\ensuremath{_1}}
-\newunicodechar{₂}{\ensuremath{_2}}
-\newunicodechar{₃}{\ensuremath{_3}}
-\newunicodechar{₄}{\ensuremath{_4}}
-\newunicodechar{₅}{\ensuremath{_5}}
-\newunicodechar{ⱼ}{\ensuremath{_j}}
-\newunicodechar{⊎}{\ensuremath{\cupdot}}
-\newunicodechar{∪}{\ensuremath{\cup}}
-\newunicodechar{ᵖ}{\ensuremath{^p}}
-\newunicodechar{≡}{\ensuremath{\equiv}}
-\newunicodechar{≢}{\ensuremath{\not\equiv}}
-\newunicodechar{∘}{\ensuremath{\circ}}
-\newunicodechar{·}{\ensuremath{\cdot}}
-\newunicodechar{ᵒ}{\ensuremath{^\circ}}
-\newunicodechar{▷}{\ensuremath{\rhd}}
-\newunicodechar{◁}{\ensuremath{\lhd}}
-\newunicodechar{∈}{\ensuremath{\in}}
-\newunicodechar{⩦}{\ensuremath{\doteq}} % !!
-\newunicodechar{∋}{\ensuremath{\ni}}
-\newunicodechar{ˢ}{\ensuremath{^s}}
-\newunicodechar{ᵈ}{\ensuremath{^d}}
-\newunicodechar{Γ}{\ensuremath{\Gamma}}
-\newunicodechar{∅}{\ensuremath{\emptyset}}
-\newunicodechar{⟨}{\ensuremath{\langle}}
-\newunicodechar{⟩}{\ensuremath{\rangle}}
-\newunicodechar{⟪}{\ensuremath{\langle\!\!\langle}}
-\newunicodechar{⟫}{\ensuremath{\rangle\!\!\rangle}}
-\newunicodechar{⟅}{\ensuremath{\Lbag}}
-\newunicodechar{⟆}{\ensuremath{\Rbag}}
-
-\newunicodechar{α}{\ensuremath{\alpha}}
-\newunicodechar{β}{\ensuremath{\beta}}
-\newunicodechar{γ}{\ensuremath{\gamma}}
-\newunicodechar{δ}{\ensuremath{\delta}}
-\newunicodechar{Δ}{\ensuremath{\Delta}}
-\newunicodechar{ξ}{\ensuremath{\xi}}
-\newunicodechar{λ}{\ensuremath{\mathnormal\lambda}}
-\newunicodechar{ƛ}{\ensuremath{\mathnormal\lambdabar}}
-\newunicodechar{Λ}{\ensuremath{\mathnormal\Lambda}}
-\newunicodechar{μ}{\ensuremath{\mu}}
-\newunicodechar{π}{\ensuremath{\pi}}
-\newunicodechar{ν}{\ensuremath{\nu}}
-\newunicodechar{ϕ}{\ensuremath{\phi}}
-\newunicodechar{ψ}{\ensuremath{\psi}}
-\newunicodechar{þ}{\ensuremath{\chi}} % !!
-\newunicodechar{σ}{\ensuremath{\sigma}}
-\newunicodechar{Σ}{\ensuremath{\Sigma}}
-\newunicodechar{τ}{\ensuremath{\tau}}
-\newunicodechar{♯}{\ensuremath{\sharp}}
-\newunicodechar{♭}{\ensuremath{\flat}}
-\newunicodechar{ℓ}{\ensuremath{\ell}}
-\newunicodechar{ₒ}{\ensuremath{_o}}
-\newunicodechar{ᵃ}{\ensuremath{^a}}
-\newunicodechar{ⁱ}{\ensuremath{^i}}
-\newunicodechar{ⱽ}{\ensuremath{^V}}
-\newunicodechar{∎}{\ensuremath{\blacksquare}} % !!
-\newunicodechar{■}{\ensuremath{\blacksquare}}
-\newunicodechar{□}{\ensuremath{\Box}}
-\newunicodechar{•}{\ensuremath{\bullet}}
-\newunicodechar{⦅}{\ensuremath{\llparenthesis}}
-\newunicodechar{⦆}{\ensuremath{\rrparenthesis}}
-\newunicodechar{⟦}{\ensuremath{\llbracket}}
-\newunicodechar{⟧}{\ensuremath{\rrbracket}}
-\newunicodechar{⦉}{\ensuremath{\llparenthesis}} % Z notation...
-\newunicodechar{⦊}{\ensuremath{\rrparenthesis}} % Z notation...
-\newunicodechar{ℰ}{\ensuremath{\mathcal{E}}}
-\newunicodechar{𝒱}{\ensuremath{\mathcal{V}}}
-\newunicodechar{𝓖}{\ensuremath{\mathcal{G}}}
-\newunicodechar{𝓔}{\ensuremath{\mathcal{E}}}
-
-%%
-%% end of the preamble, start of the body of the document source.
-\begin{document}
-
-%%
-%% The "title" command has an optional parameter,
-%% allowing the author to define a "short title" to be used in page headers.
-\title{A Step-Indexed Logic in Agda}
-
-%%
-%% The "author" command and its associated commands are used to define
-%% the authors and their affiliations.
-%% Of note is the shared affiliation of the first two authors, and the
-%% "authornote" and "authornotemark" commands
-%% used to denote shared contribution to the research.
-\author{Jeremy G. Siek}
-%\authornote{Both authors contributed equally to this research.}
-\email{jsiek@indiana.edu}
-\orcid{0000-0002-9894-4856}
-%% \author{G.K.M. Tobin}
-%% \authornotemark[1]
-%% \email{webmaster@marysville-ohio.com}
-\affiliation{%
-  \institution{Indiana University}
-  \streetaddress{Luddy Hall, 700 N. Woodlawn Avenue}
-  \city{Bloomington}
-  \state{IN}
-  \country{USA}
-  \postcode{47408}
-}
-
-\author{Philip Wadler}
-\affiliation{%
-  \institution{University of Edinburgh}
-  %% \streetaddress{1 Th{\o}rv{\"a}ld Circle}
-  \city{Edinburgh}
-  \country{Scotland}}
-%% \email{larst@affiliation.org}
-
-\author{Peter Thiemann}
-\affiliation{%
-  \institution{University of Freiburg}
-  \city{Freiburg}
-  \country{Germany}
-}
-
-%%
-%% By default, the full list of authors will be used in the page
-%% headers. Often, this list is too long, and will overlap
-%% other information printed in the page headers. This command allows
-%% the author to define a more concise list
-%% of authors' names for this purpose.
-%\renewcommand{\shortauthors}{Trovato et al.}
-
-%%
-%% The abstract is a short summary of the work to be presented in the
-%% article.
-\begin{abstract}
-  Step-indexed logical relations are an important tool in the
-  metatheory of programming
-  languages~\citep{Appel:2001aa,Ahmed:2006aa}, and their application
-  can be streamlined by use of step-indexed modal logics such as
-  LSLR~\citep{Dreyer:2011wl}.  This paper describes an embedding of a
-  step-indexed modal logic in Agda and its application to a semantic
-  type safety proof for the Simply-Typed Lambda Calculus extended with
-  recursive functions.
-\end{abstract}
-
-%%
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-%% Please copy and paste the code instead of the example below.
-%%
-%\begin{CCSXML}
-%\end{CCSXML}
-
-%% \ccsdesc[500]{Computer systems organization~Embedded systems}
-%% \ccsdesc[300]{Computer systems organization~Redundancy}
-%% \ccsdesc{Computer systems organization~Robotics}
-%% \ccsdesc[100]{Networks~Network reliability}
-
-%%
-%% Keywords. The author(s) should pick words that accurately describe
-%% the work being presented. Separate the keywords with commas.
-\keywords{step-indexed logical relations, modal logic}
-
-%% \received{20 February 2007}
-%% \received[revised]{12 March 2009}
-%% \received[accepted]{5 June 2009}
-
-%%
-%% This command processes the author and affiliation and title
-%% information and builds the first part of the formatted document.
-\maketitle
-
-%TODO: Road Map/Outline of what's in the paper
-
-\clearpage
-
-\tableofcontents
-
-\section{Introduction}
-\label{sec:intro}
-
-Logical relations are a useful technique for proving properties of
-programming languages, from type safety~\citep{Timany:2022aa} to
-relational parametricity~\citep{REYNOLDS83}, and
-noninterference~\citep{heintze1998slam}.  Initially, logical relations
-were applied to terminating languages such as the Simply Typed Lambda
-Calculus~\citep{Tait:1967aa} and System F~\citep{GIRARD72}.
-\citet{Appel:2001aa} introduced step-indexed logical relations to
-handle languages with recursive types.  \citet{Dreyer:2011wl} designed
-the LSLR logic to streamline the reasoning about step indexed logical
-relations.  This work eventually led to Iris~\citep{JUNG:2018aa}, a
-system that supports a combination of step indexing and separation
-logic within the Coq proof
-assistant~\citep{The-Coq-Development-Team:2004kf,Huet:2016aa}.
-
-The authors are interested in using step-indexed logical relations for
-purposes of developing mechanized proofs in Agda of the Gradual
-Guarantee~\citep{Siek:2015ac} and Parametricity~\citep{REYNOLDS74C}
-for gradually-typed polymorphic
-languages~\citep{Ahmed:2011fk,Ahmed:2017aa,Igarashi:2017aa,New:2019ab,Labrada:2020tk}. One
-option would be to use Guarded Cubical Agda~\citep{Veltri:2020aa}, but
-both Cubical Agda~\citep{Vezzosi:2019aa} and its extension with
-guarded fixpoints are experimental features of Agda that may not be
-supported indefinitely.  We would prefer to have a step-indexed logic
-that is embedded in (vanilla) Agda.
-
-This paper fills that need by showing how to build a step-indexed
-logic in Agda, a system that we abbreviate as SIL.  This paper is a
-literate Agda file; all of the definitions and proofs have been
-checked with Agda 2.6.2.2. 
-%
-Most of the challenge in defining a step-indexed logic is in the
-definition of recursive predicates and their fixpoint theorem. The
-step-indexed models of \citet{Appel:2001aa} rely on an approximation
-operator, here written as $↓_n$, to define the semantic criteria for
-valid recursive definitions.  For example, they define a nonexpansive
-function to be one that satisfies
-\[
-  ↓_n f(x) = ↓_n f(↓_n x)
-\]
-The step-indexed models of \citet{JUNG:2018aa} instead rely on
-a step-indexed notion of equality. For example, they define a
-nonexpansive function as one that satisfies
-\[
-  x \stackrel{n}{=} y \text{ implies } f(x) \stackrel{n}{=} f(y)
-\]
-We are not aware of any inherent reason to prefer one approach to the
-other, so we choose the approach of \citet{Appel:2001aa} because they
-provide perspicuous proofs of their fixpoint theorem.  So our approach
-to SIL can be seen as an implementation of the LSLR logic of
-\citet{Dreyer:2011wl} using the foundations of
-\citet{Appel:2001aa}. In particular, SIL defines the recursive
-predicate of SIL using function iteration, like this:
-\[
-  ⟦ μ F ⟧ \, x \, n = F^{n\plus 1}\, \top \, x \, n
-\]
-whereas LSLR uses substitution in the semantics of recursive types:
-\[
-   ⟦ μ r. R ⟧ \, x \, n = ⟦ R[μr.R/r] ⟧\,x\,n
-\]
-
-However, adapting the work of \citet{Appel:2001aa} to our setting was
-nontrivial because (1) their model is for recursive types, not a modal
-logic, and (2) their proofs were incomplete (which we learned part way
-through our development of SIL). To enable the nesting of recursive
-types inside other recursive types, one must prove that the $μ$ type
-constructor is nonexpansive and well founded, but \citet{Appel:2001aa}
-does not prove this (they prove those properties for the other type
-constructors).  We discuss this problem and its remedy in sections
-\ref{sec:appel-mcallester} and \ref{sec:open-propositions}.
-
-We organize this paper as follows. We review Appel an McAllester's
-work in Section~\ref{sec:appel-mcallester}, then introduce our
-representation of step-indexed propositions and predicates in Agda in
-Sections \ref{sec:propositions} and \ref{sec:predicates}.  We
-formalize several important concepts in
-Section~\ref{sec:fun-approx-iter}: functionals, approximation, and
-iteration, and prove some basic results about them.  We construct all
-of the logical connectives of SIL in
-Section~\ref{sec:open-propositions}, including the definition of
-recursive predicates and their fixpoint theorem.  The proof theory of
-SIL is developed in Section~\ref{sec:proofs}, where we construct the
-proof rules of this modal logic.  To evaluate the utility of SIL,
-Section~\ref{sec:STLC} presents a case study of proving semantic type
-safety for the Simply Typed Lambda Calculus extended with recursive
-functions. We find that SIL streamlines the definition of logical
-relations and is helpful in many of the proofs, but there is room for
-improvement regarding the ergonomics of interactive proof development
-using SIL.  Limitations in Agda's inference algorithm sometimes leads
-to verbosity in SIL proofs.
-
-%% Our goal is to define an operator for recursive predicates and relations
-%% with syntax that is something like $μᵒ x. R$, where $x$ is the name of the
-%% recursive relation and $R$ is the definition of the relation, which
-%% can refer to $x$. We shall prove a fixpoint theorem which states that
-%% the recursive predicate is equal to its unfolding, something like the
-%% following.
-%% \[
-%%   (μᵒ x. R) \app δ \app a ≡ᵒ R \app δ(x:= μᵒ x. R) \app a
-%% \]
-%% where $δ$ is an environment mapping variables to predicates.
-
 \section{Review of Appel and McAllester's Fixpoint Theorem}
@@
 the logic.
-
-\input{latex/cpp2024/StepIndexedLogic.tex}
-
-\input{latex/cpp2024/STLC.tex}
-\input{latex/cpp2024/STLCBind.tex}
-\input{latex/cpp2024/STLCSafe.tex}
-
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{all}
-
-\section*{Appendix}
-
-\input{latex/cpp2024/STLCDeterministic.tex}
-
-\end{document}
-\endinput

### 2023-09-14 papers/July2023/paper.tex
check
equiv, subst, and fixpoint

@@
 \newunicodechar{ᵒ}{\ensuremath{^\circ}}
+\newunicodechar{″}{\ensuremath{''}}
 \newunicodechar{▷}{\ensuremath{\rhd}}
@@
 \newunicodechar{þ}{\ensuremath{\chi}} % !!
+\newunicodechar{χ}{\ensuremath{\chi}}
 \newunicodechar{σ}{\ensuremath{\sigma}}
@@
 \newunicodechar{ₒ}{\ensuremath{_o}}
+\newunicodechar{ₛ}{\ensuremath{_s}}
+\newunicodechar{ₚ}{\ensuremath{_p}}
+\newunicodechar{ₐ}{\ensuremath{_a}}
 \newunicodechar{ᵃ}{\ensuremath{^a}}
@@
 \newunicodechar{𝓔}{\ensuremath{\mathcal{E}}}
+\newunicodechar{⨟}{\ensuremath{\fatsemi}}
+\newunicodechar{≐}{\ensuremath{\doteq}}
+\newunicodechar{⊂}{\ensuremath{\subset}}
+\newunicodechar{⌊}{\ensuremath{\lfloor}}
+\newunicodechar{⌋}{\ensuremath{\rfloor}}
+\newunicodechar{ᵇ}{\ensuremath{^b}}
+\newunicodechar{ᵏ}{\ensuremath{^k}}
+\newunicodechar{◇}{\ensuremath{\diamond}}
+\newunicodechar{ˡ}{\ensuremath{^l}}
+\newunicodechar{⁰}{\ensuremath{^0}}
 
@@
 
+%
+%  step-indexed logics 
+%
+%
+%
+
 %%
@@
 
-\section{Introduction}
+\section{Overview}
 \label{sec:intro}
@@
 for gradually-typed polymorphic
-languages~\citep{Ahmed:2011fk,Ahmed:2017aa,Igarashi:2017aa,New:2019ab,Labrada:2020tk}. One
-option would be to use Guarded Cubical Agda~\citep{Veltri:2020aa}, but
-both Cubical Agda~\citep{Vezzosi:2019aa} and its extension with
-guarded fixpoints are experimental features of Agda that may not be
-supported indefinitely.  We would prefer to have a step-indexed logic
-that is embedded in (vanilla) Agda.
+languages~\citep{Ahmed:2011fk,Ahmed:2017aa,Igarashi:2017aa,New:2019ab,Labrada:2020tk}.
+One option is to use Guarded Cubical Agda~\citep{Veltri:2020aa}, an
+experimental extension to Agda, and another option is to develop an
+embedding of a step-indexed logic in (vanilla) Agda. The authors are
+interested in both approaches, as there are tradeoffs that should be
+evaluated. For example, using Cubical Agda comes with the down-side of
+losing Uniqueness of Identity Proofs (UIP) but on the other hand
+Guarded Cubical Agda provides a more ergonomic user-interface for
+step-indexed reasoning than an embedded logic. The authors chose to
+first investigate the embedding approach, which we describe in this
+paper.
 
-This paper fills that need by showing how to build a step-indexed
-logic in Agda, a system that we abbreviate as SIL.  This paper is a
-literate Agda file; all of the definitions and proofs have been
-checked with Agda 2.6.2.2.
+This paper shows how to embed a step-indexed logic in Agda, developing
+a system that we abbreviate as SIL. This paper is a literate Agda
+file; all of the definitions and proofs have been checked with Agda
+2.6.2.2.
 
@@
 
-\input{latex/July2023/STLC.tex}
-\input{latex/July2023/STLCBind.tex}
-\input{latex/July2023/STLCSafe.tex}
+\input{latex/July2023/STLC2.tex}
+\input{latex/July2023/LogRel2.tex}
+\input{latex/July2023/STLCDeterministic2.tex}
+\input{latex/July2023/STLCBind2.tex}
+\input{latex/July2023/STLCSafe2.tex}
+\input{latex/July2023/STLCTypeSafe.tex}
 
@@
 
-\input{latex/July2023/StepIndexedLogic.tex}
+\input{../../src/latex/StepIndexedLogic2.tex}
+% \input{latex/July2023/SILReference.tex}
 
@@
 
-\input{latex/July2023/STLCDeterministic.tex}
-
 \end{document}

### 2023-09-22 papers/July2023/paper.tex
updates to section 3

@@
 Guarded Cubical Agda provides a more ergonomic user-interface for
-step-indexed reasoning than an embedded logic. The authors chose to
-first investigate the embedding approach, which we describe in this
-paper.
+step-indexed reasoning than what's possible in an embedded logic. The
+authors chose to first investigate the embedding approach.
 
@@
 2.6.2.2.
+
+We describe SIL from the user perspective in
+§\ref{sec:SIL-intro}, including what kinnds of formulas can be
+written in SIL and how to conduct proofs in SIL. As a step-indexed
+logic, SIL supports user-defined recursive predicates and a ``later''
+operator.
 
@@
 
-UNDER CONSTRUCTION
-
-To evaluate the utility of SIL, Section~\ref{sec:STLC} presents a case
+To evaluate the utility of SIL, §\ref{sec:STLC} presents a case
 study of proving semantic type safety for the Simply Typed Lambda
@@
 
+TODO: roadmap for the implementation of SIL
+
 \input{latex/July2023/SILIntro.tex}
@@
 \input{latex/July2023/LogRel2.tex}
-\input{latex/July2023/STLCDeterministic2.tex}
 \input{latex/July2023/STLCBind2.tex}
@@
 We organize the rest of the article as follows. We review Appel an
-McAllester's work in Section~\ref{sec:appel-mcallester}, then
+McAllester's work in §\ref{sec:appel-mcallester}, then
 introduce our representation of step-indexed propositions and
@@
 \ref{sec:predicates}.  We formalize several important concepts in
-Section~\ref{sec:fun-approx-iter}: functionals, approximation, and
+§\ref{sec:fun-approx-iter}: functionals, approximation, and
 iteration, and prove some basic results about them.  We construct all
 of the logical connectives of SIL in
-Section~\ref{sec:open-propositions}, including the definition of
+§\ref{sec:open-propositions}, including the definition of
 recursive predicates and their fixpoint theorem.  The proof theory of
-SIL is developed in Section~\ref{sec:proofs}, where we construct the
+SIL is developed in §\ref{sec:proofs}, where we construct the
 proof rules of this modal logic.
@@
 
+\input{latex/July2023/STLCDeterministic2.tex}
+
 \end{document}

### 2023-09-26 papers/July2023/notes.tex
a little progress

--- 2023-09-22 papers/July2023/paper.tex
@@
-%%
-%% This is file `sample-acmsmall.tex',
-%% generated with the docstrip utility.
-%%
-%% The original source files were:
-%%
-%% samples.dtx  (with options: `acmsmall')
-%% 
-%% IMPORTANT NOTICE:
-%% 
-%% For the copyright see the source file.
-%% 
-%% Any modified versions of this file must be renamed
-%% with new filenames distinct from sample-acmsmall.tex.
-%% 
-%% For distribution of the original source see the terms
-%% for copying and modification in the file samples.dtx.
-%% 
-%% This generated file may be distributed as long as the
-%% original source files, as listed above, are part of the
-%% same distribution. (The sources need not necessarily be
-%% in the same archive or directory.)
-%%
-%%
-%% Commands for TeXCount
-%TC:macro \cite [option:text,text]
-%TC:macro \citep [option:text,text]
-%TC:macro \citet [option:text,text]
-%TC:envir table 0 1
-%TC:envir table* 0 1
-%TC:envir tabular [ignore] word
-%TC:envir displaymath 0 word
-%TC:envir math 0 word
-%TC:envir comment 0 0
-%%
-%%
-%% The first command in your LaTeX source must be the \documentclass
-%% command.
-%%
-%% For submission and review of your manuscript please change the
-%% command to \documentclass[manuscript, screen, review]{acmart}.
-%%
-%% When submitting camera ready or to TAPS, please change the command
-%% to \documentclass[sigconf]{acmart} or whichever template is required
-%% for your publication.
-%%
-%%
-\documentclass[acmsmall]{acmart}
-\usepackage{semantic}
-\usepackage{stmaryrd}
-\usepackage{agda}
-\usepackage{MnSymbol}
-
-%%
-%% \BibTeX command to typeset BibTeX logo in the docs
-\AtBeginDocument{%
-  \providecommand\BibTeX{{%
-    Bib\TeX}}}
-
-%% Rights management information.  This information is sent to you
-%% when you complete the rights form.  These commands have SAMPLE
-%% values in them; it is your responsibility as an author to replace
-%% the commands and values with those provided to you when you
-%% complete the rights form.
-\setcopyright{acmcopyright}
-\copyrightyear{2023}
-\acmYear{2023}
-\acmDOI{XXXXXXX.XXXXXXX}
-
-%%
-%% These commands are for a JOURNAL article.
-%% \acmJournal{JACM}
-%% \acmVolume{37}
-%% \acmNumber{4}
-%% \acmArticle{111}
-%% \acmMonth{8}
-
-%%
-%% Submission ID.
-%% Use this when submitting an article to a sponsored event. You'll
-%% receive a unique submission ID from the organizers
-%% of the event, and this ID should be used as the parameter to this command.
-%%\acmSubmissionID{123-A56-BU3}
-
-%%
-%% For managing citations, it is recommended to use bibliography
-%% files in BibTeX format.
-%%
-%% You can then either use BibTeX with the ACM-Reference-Format style,
-%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
-%% support for advanced citation of software artefact from the
-%% biblatex-software package, also separately available on CTAN.
-%%
-%% Look at the sample-*-biblatex.tex files for templates showcasing
-%% the biblatex styles.
-%%
-
-%%
-%% The majority of ACM publications use numbered citations and
-%% references.  The command \citestyle{authoryear} switches to the
-%% "author year" style.
-%%
-%% If you are preparing content for an event
-%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
-%% citations and references.
-%% Uncommenting
-%% the next command will enable that style.
-\citestyle{acmauthoryear}
-
-\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
-\newcommand{\semV}[1]{\mathcal{V}\sem{#1}}
-\newcommand{\semE}[1]{\mathcal{E}\sem{#1}}
-\newcommand{\ba}{\begin{array}}
-\newcommand{\ea}{\end{array}}
-\newenvironment{stack}{\ba{@{}l@{}}}{\ea}
-\newenvironment{branch}{\left\{\ba{@{}l@{\qquad}l@{}}}{\ea\right\}}
-\newenvironment{syntax}{\[\ba{l@{\;\;}lcl}}{\ea\]}
-\newcommand{\dotspace}{.\,}
-\newcommand{\key}[1]{\ensuremath{\mathtt{#1}}}
-\newcommand{\dyn}{\star}
-\newcommand{\Dyn}{\ensuremath{\dyn}}
-\newcommand{\Int}{\key{int}}
-\newcommand{\Bool}{\key{bool}}
-\newcommand{\lam}[1]{\lambda #1 \dotspace}
-\newcommand{\app}{\,}
-\newcommand{\reduce}{\longrightarrow}
-\newcommand{\inj}[2]{#1 \,!\, #2}
-\newcommand{\proj}[2]{#1 \,?\, #2}
-% \newcommand{\kapprox}[2]{\mathrm{approx}(#1, #2)}
-\newcommand{\kapprox}[2]{↓_{#1}\, #2}
-\newcommand{\varnow}{\mathsf{var}\mbox{-}\mathsf{now}}
-\newcommand{\laters}{\mathsf{laters}}
-\newcommand{\Now}{\mathsf{Now}}
-\newcommand{\Later}{\mathsf{Later}}
-\newcommand{\zero}{\mathsf{zero}}
-\newcommand{\suc}[1]{\mathsf{suc}\app#1}
-\newcommand{\eff}[1]{\Lbag#1\Rbag}
-\newcommand{\AgdaMissingDefinition}[1]{#1}
-
-\usepackage{newunicodechar}
-\newunicodechar{←}{\ensuremath{\mathnormal\from}}
-\newunicodechar{→}{\ensuremath{\mathnormal\to}}
-\newunicodechar{↓}{\ensuremath{\mathnormal\downarrow}}
-\newunicodechar{⇑}{\ensuremath{\mathnormal\Uparrow}}
-\newunicodechar{⇓}{\ensuremath{\mathnormal\Downarrow}}
-\newunicodechar{⇒}{\ensuremath{\Rightarrow}}
-\newunicodechar{⇔}{\ensuremath{\iff}}
-\newunicodechar{↠}{\ensuremath{\rightarrow^{\ast}}}
-\newunicodechar{∀}{\ensuremath{\mathnormal\forall}}
-\newunicodechar{∃}{\ensuremath{\exists}}
-\newunicodechar{Π}{\ensuremath{\Pi}}
-\newunicodechar{⊢}{\ensuremath{\vdash}}
-\newunicodechar{⊣}{\ensuremath{\dashv}}
-\newunicodechar{⊧}{\ensuremath{\models}}
-\newunicodechar{⊨}{\ensuremath{\models}}
-\newunicodechar{↦}{\ensuremath{\mapsto}}
-\newunicodechar{⊥}{\ensuremath{\mathnormal\bot}}
-\newunicodechar{⊤}{\ensuremath{\mathnormal\top}}
-\newunicodechar{∷}{\ensuremath{\mathnormal::}}
-\newunicodechar{⦂}{\ensuremath{:}}  % !!
-\newunicodechar{ℕ}{\ensuremath{\mathbb N}}
-\newunicodechar{ℤ}{\ensuremath{\mathbb Z}}
-\newunicodechar{𝒫}{\ensuremath{\mathcal P}}
-\newunicodechar{≤}{\ensuremath{\leq}}
-\newunicodechar{×}{\ensuremath{\times}}
-\newunicodechar{∸}{\ensuremath{\minus}}
-\newunicodechar{′}{\ensuremath{'}}
-\newunicodechar{₁}{\ensuremath{_1}}
-\newunicodechar{₂}{\ensuremath{_2}}
-\newunicodechar{₃}{\ensuremath{_3}}
-\newunicodechar{₄}{\ensuremath{_4}}
-\newunicodechar{₅}{\ensuremath{_5}}
-\newunicodechar{ⱼ}{\ensuremath{_j}}
-\newunicodechar{⊎}{\ensuremath{\cupdot}}
-\newunicodechar{∪}{\ensuremath{\cup}}
-\newunicodechar{ᵖ}{\ensuremath{^p}}
-\newunicodechar{≡}{\ensuremath{\equiv}}
-\newunicodechar{≢}{\ensuremath{\not\equiv}}
-\newunicodechar{∘}{\ensuremath{\circ}}
-\newunicodechar{·}{\ensuremath{\cdot}}
-\newunicodechar{ᵒ}{\ensuremath{^\circ}}
-\newunicodechar{″}{\ensuremath{''}}
-\newunicodechar{▷}{\ensuremath{\rhd}}
-\newunicodechar{◁}{\ensuremath{\lhd}}
-\newunicodechar{∈}{\ensuremath{\in}}
-\newunicodechar{⩦}{\ensuremath{\doteq}} % !!
-\newunicodechar{∋}{\ensuremath{\ni}}
-\newunicodechar{ˢ}{\ensuremath{^s}}
-\newunicodechar{ᵈ}{\ensuremath{^d}}
-\newunicodechar{Γ}{\ensuremath{\Gamma}}
-\newunicodechar{∅}{\ensuremath{\emptyset}}
-\newunicodechar{⟨}{\ensuremath{\langle}}
-\newunicodechar{⟩}{\ensuremath{\rangle}}
-\newunicodechar{⟪}{\ensuremath{\langle\!\!\langle}}
-\newunicodechar{⟫}{\ensuremath{\rangle\!\!\rangle}}
-\newunicodechar{⟅}{\ensuremath{\Lbag}}
-\newunicodechar{⟆}{\ensuremath{\Rbag}}
-
-\newunicodechar{α}{\ensuremath{\alpha}}
-\newunicodechar{β}{\ensuremath{\beta}}
-\newunicodechar{γ}{\ensuremath{\gamma}}
-\newunicodechar{δ}{\ensuremath{\delta}}
-\newunicodechar{Δ}{\ensuremath{\Delta}}
-\newunicodechar{ξ}{\ensuremath{\xi}}
-\newunicodechar{λ}{\ensuremath{\mathnormal\lambda}}
-\newunicodechar{ƛ}{\ensuremath{\mathnormal\lambdabar}}
-\newunicodechar{Λ}{\ensuremath{\mathnormal\Lambda}}
-\newunicodechar{μ}{\ensuremath{\mu}}
-\newunicodechar{π}{\ensuremath{\pi}}
-\newunicodechar{ν}{\ensuremath{\nu}}
-\newunicodechar{ϕ}{\ensuremath{\phi}}
-\newunicodechar{ψ}{\ensuremath{\psi}}
-\newunicodechar{þ}{\ensuremath{\chi}} % !!
-\newunicodechar{χ}{\ensuremath{\chi}}
-\newunicodechar{σ}{\ensuremath{\sigma}}
-\newunicodechar{Σ}{\ensuremath{\Sigma}}
-\newunicodechar{τ}{\ensuremath{\tau}}
-\newunicodechar{♯}{\ensuremath{\sharp}}
-\newunicodechar{♭}{\ensuremath{\flat}}
-\newunicodechar{ℓ}{\ensuremath{\ell}}
-\newunicodechar{ₒ}{\ensuremath{_o}}
-\newunicodechar{ₛ}{\ensuremath{_s}}
-\newunicodechar{ₚ}{\ensuremath{_p}}
-\newunicodechar{ₐ}{\ensuremath{_a}}
-\newunicodechar{ᵃ}{\ensuremath{^a}}
-\newunicodechar{ⁱ}{\ensuremath{^i}}
-\newunicodechar{ⱽ}{\ensuremath{^V}}
-\newunicodechar{∎}{\ensuremath{\blacksquare}} % !!
-\newunicodechar{■}{\ensuremath{\blacksquare}}
-\newunicodechar{□}{\ensuremath{\Box}}
-\newunicodechar{•}{\ensuremath{\bullet}}
-\newunicodechar{⦅}{\ensuremath{\llparenthesis}}
-\newunicodechar{⦆}{\ensuremath{\rrparenthesis}}
-\newunicodechar{⟦}{\ensuremath{\llbracket}}
-\newunicodechar{⟧}{\ensuremath{\rrbracket}}
-\newunicodechar{⦉}{\ensuremath{\llparenthesis}} % Z notation...
-\newunicodechar{⦊}{\ensuremath{\rrparenthesis}} % Z notation...
-\newunicodechar{ℰ}{\ensuremath{\mathcal{E}}}
-\newunicodechar{𝒱}{\ensuremath{\mathcal{V}}}
-\newunicodechar{𝓖}{\ensuremath{\mathcal{G}}}
-\newunicodechar{𝓔}{\ensuremath{\mathcal{E}}}
-\newunicodechar{⨟}{\ensuremath{\fatsemi}}
-\newunicodechar{≐}{\ensuremath{\doteq}}
-\newunicodechar{⊂}{\ensuremath{\subset}}
-\newunicodechar{⌊}{\ensuremath{\lfloor}}
-\newunicodechar{⌋}{\ensuremath{\rfloor}}
-\newunicodechar{ᵇ}{\ensuremath{^b}}
-\newunicodechar{ᵏ}{\ensuremath{^k}}
-\newunicodechar{◇}{\ensuremath{\diamond}}
-\newunicodechar{ˡ}{\ensuremath{^l}}
-\newunicodechar{⁰}{\ensuremath{^0}}
-
-%%
-%% end of the preamble, start of the body of the document source.
-\begin{document}
-
-%%
-%% The "title" command has an optional parameter,
-%% allowing the author to define a "short title" to be used in page headers.
-\title{A Step-Indexed Logic in Agda}
-
-%%
-%% The "author" command and its associated commands are used to define
-%% the authors and their affiliations.
-%% Of note is the shared affiliation of the first two authors, and the
-%% "authornote" and "authornotemark" commands
-%% used to denote shared contribution to the research.
-\author{Jeremy G. Siek}
-%\authornote{Both authors contributed equally to this research.}
-\email{jsiek@indiana.edu}
-\orcid{0000-0002-9894-4856}
-%% \author{G.K.M. Tobin}
-%% \authornotemark[1]
-%% \email{webmaster@marysville-ohio.com}
-\affiliation{%
-  \institution{Indiana University}
-  \streetaddress{Luddy Hall, 700 N. Woodlawn Avenue}
-  \city{Bloomington}
-  \state{IN}
-  \country{USA}
-  \postcode{47408}
-}
-
-\author{Philip Wadler}
-\affiliation{%
-  \institution{University of Edinburgh}
-  %% \streetaddress{1 Th{\o}rv{\"a}ld Circle}
-  \city{Edinburgh}
-  \country{Scotland}}
-%% \email{larst@affiliation.org}
-
-\author{Peter Thiemann}
-\affiliation{%
-  \institution{University of Freiburg}
-  \city{Freiburg}
-  \country{Germany}
-}
-
-%%
-%% By default, the full list of authors will be used in the page
-%% headers. Often, this list is too long, and will overlap
-%% other information printed in the page headers. This command allows
-%% the author to define a more concise list
-%% of authors' names for this purpose.
-%\renewcommand{\shortauthors}{Trovato et al.}
-
-%
-%  step-indexed logics 
-%
-%
-%
-
-%%
-%% The abstract is a short summary of the work to be presented in the
-%% article.
-\begin{abstract}
-  Step-indexed logical relations are an important tool in the
-  metatheory of programming
-  languages~\citep{Appel:2001aa,Ahmed:2006aa}, and their application
-  can be streamlined by use of step-indexed modal logics such as
-  LSLR~\citep{Dreyer:2011wl}.  This paper describes a mechanization of
-  a step-indexed modal logic in the Agda proof assistant and its
-  application to a semantic type safety proof for the Simply-Typed
-  Lambda Calculus extended with recursive functions. The step-indexed
-  logic supports guarded recursive predicates via the later modality
-  and provides a fixpoint theorem adapted from \citet{Appel:2001aa}.
-\end{abstract}
-
-%%
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-%% Please copy and paste the code instead of the example below.
-%%
-%\begin{CCSXML}
-%\end{CCSXML}
-
-%% \ccsdesc[500]{Computer systems organization~Embedded systems}
-%% \ccsdesc[300]{Computer systems organization~Redundancy}
-%% \ccsdesc{Computer systems organization~Robotics}
-%% \ccsdesc[100]{Networks~Network reliability}
-
-%%
-%% Keywords. The author(s) should pick words that accurately describe
-%% the work being presented. Separate the keywords with commas.
-\keywords{step-indexed logical relations, modal logic}
-
-%% \received{20 February 2007}
-%% \received[revised]{12 March 2009}
-%% \received[accepted]{5 June 2009}
-
-%%
-%% This command processes the author and affiliation and title
-%% information and builds the first part of the formatted document.
-\maketitle
-
-%TODO: Road Map/Outline of what's in the paper
-
-\clearpage
-
-\tableofcontents
-
-\section{Overview}
-\label{sec:intro}
-
-Logical relations are a useful technique for proving properties of
-programming languages, from type safety~\citep{Timany:2022aa} to
-relational parametricity~\citep{REYNOLDS83}, and
-noninterference~\citep{heintze1998slam}.  Initially, logical relations
-were applied to terminating languages such as the Simply Typed Lambda
-Calculus~\citep{Tait:1967aa} and System F~\citep{GIRARD72}.
-\citet{Appel:2001aa} introduced step-indexed logical relations to
-handle languages with recursive types.  \citet{Dreyer:2011wl} designed
-the LSLR logic to streamline the reasoning about step indexed logical
-relations.  This work eventually led to Iris~\citep{JUNG:2018aa}, a
-system that supports a combination of step indexing and separation
-logic within the Coq proof
-assistant~\citep{The-Coq-Development-Team:2004kf,Huet:2016aa}.
-
-The authors are interested in using step-indexed logical relations for
-purposes of developing mechanized proofs in Agda of the Gradual
-Guarantee~\citep{Siek:2015ac} and Parametricity~\citep{REYNOLDS74C}
-for gradually-typed polymorphic
-languages~\citep{Ahmed:2011fk,Ahmed:2017aa,Igarashi:2017aa,New:2019ab,Labrada:2020tk}.
-One option is to use Guarded Cubical Agda~\citep{Veltri:2020aa}, an
-experimental extension to Agda, and another option is to develop an
-embedding of a step-indexed logic in (vanilla) Agda. The authors are
-interested in both approaches, as there are tradeoffs that should be
-evaluated. For example, using Cubical Agda comes with the down-side of
-losing Uniqueness of Identity Proofs (UIP) but on the other hand
-Guarded Cubical Agda provides a more ergonomic user-interface for
-step-indexed reasoning than what's possible in an embedded logic. The
-authors chose to first investigate the embedding approach.
-
-This paper shows how to embed a step-indexed logic in Agda, developing
-a system that we abbreviate as SIL. This paper is a literate Agda
-file; all of the definitions and proofs have been checked with Agda
-2.6.2.2.
-
-We describe SIL from the user perspective in
-§\ref{sec:SIL-intro}, including what kinnds of formulas can be
-written in SIL and how to conduct proofs in SIL. As a step-indexed
-logic, SIL supports user-defined recursive predicates and a ``later''
-operator.
-
-%% Our goal is to define an operator for recursive predicates and relations
-%% with syntax that is something like $μᵒ x. R$, where $x$ is the name of the
-%% recursive relation and $R$ is the definition of the relation, which
-%% can refer to $x$. We shall prove a fixpoint theorem which states that
-%% the recursive predicate is equal to its unfolding, something like the
-%% following.
-%% \[
-%%   (μᵒ x. R) \app δ \app a ≡ᵒ R \app δ(x:= μᵒ x. R) \app a
-%% \]
-%% where $δ$ is an environment mapping variables to predicates.
-
-To evaluate the utility of SIL, §\ref{sec:STLC} presents a case
-study of proving semantic type safety for the Simply Typed Lambda
-Calculus extended with recursive functions. We find that SIL
-streamlines the definition of logical relations and is helpful in many
-of the proofs, but there is room for improvement regarding the
-ergonomics of interactive proof development using SIL.  Limitations in
-Agda's inference algorithm sometimes leads to verbosity in SIL proofs.
-
-TODO: roadmap for the implementation of SIL
-
-\input{latex/July2023/SILIntro.tex}
-
-\input{latex/July2023/STLC2.tex}
-\input{latex/July2023/LogRel2.tex}
-\input{latex/July2023/STLCBind2.tex}
-\input{latex/July2023/STLCSafe2.tex}
-\input{latex/July2023/STLCTypeSafe.tex}
-
-\section{Constructing a Step-Indexed Logic}
-
 Most of the challenge in defining a step-indexed logic is in the
@@
 proof rules of this modal logic.
-
-\input{review-appel.tex}
-
-\input{../../src/latex/StepIndexedLogic2.tex}
-% \input{latex/July2023/SILReference.tex}
-
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{all}
-
-\appendix
-\addcontentsline{toc}{section}{Appendix}
-\section*{Appendix}
-
-\input{latex/July2023/STLCDeterministic2.tex}
-
-\end{document}
-\endinput

### 2023-09-26 papers/July2023/paper.tex
a little progress

@@
 \newunicodechar{∪}{\ensuremath{\cup}}
+\newunicodechar{⊔}{\ensuremath{\sqcup}}
+
 \newunicodechar{ᵖ}{\ensuremath{^p}}
@@
 
-Most of the challenge in defining a step-indexed logic is in the
-definition of recursive predicates and their fixpoint theorem. The
-step-indexed models of \citet{Appel:2001aa} rely on an approximation
-operator, here written as $↓_n$, to define the semantic criteria for
-valid recursive definitions.  For example, they define a nonexpansive
-function to be one that satisfies
-\[
-  ↓_n f(x) = ↓_n f(↓_n x)
-\]
-The step-indexed models of \citet{JUNG:2018aa} instead rely on
-a step-indexed notion of equality. For example, they define a
-nonexpansive function as one that satisfies
-\[
-  x \stackrel{n}{=} y \text{ implies } f(x) \stackrel{n}{=} f(y)
-\]
-We are not aware of any inherent reason to prefer one approach to the
-other, so we choose the approach of \citet{Appel:2001aa} because they
-provide perspicuous proofs of their fixpoint theorem.  So our approach
-to SIL can be seen as an implementation of the LSLR logic of
-\citet{Dreyer:2011wl} using the foundations of
-\citet{Appel:2001aa}. In particular, SIL defines the recursive
-predicate of SIL using function iteration, like this:
-\[
-  ⟦ μ F ⟧ \, x \, n = F^{n\plus 1}\, \top \, x \, n
-\]
-whereas LSLR uses substitution in the semantics of recursive types:
-\[
-   ⟦ μ r. R ⟧ \, x \, n = ⟦ R[μr.R/r] ⟧\,x\,n
-\]
+\input{../../src/latex/RawSetO.tex}
 
-However, adapting the work of \citet{Appel:2001aa} to our setting was
-nontrivial because (1) their model is for recursive types, not a modal
-logic, and (2) their proofs were incomplete (which we learned part way
-through our development of SIL). To enable the nesting of recursive
-types inside other recursive types, one must prove that the $μ$ type
-constructor is nonexpansive and well founded, but \citet{Appel:2001aa}
-does not prove this (they prove those properties for the other type
-constructors).  We discuss this problem and its remedy in sections
-\ref{sec:appel-mcallester} and \ref{sec:open-propositions}.
+Unfortunately, the \textsf{fixpoint}ᵒ theorem that we introduced in
+\S~\ref{sec:formula-equivalence} cannot be proved for arbitrary
+functions from ℕ to \textsf{Prop}. As usual, one needs to place some
+restrictions on the function space. We place three restrictions on
+these functions: (1) they must be downward closed
 
-We organize the rest of the article as follows. We review Appel an
-McAllester's work in §\ref{sec:appel-mcallester}, then
-introduce our representation of step-indexed propositions and
-predicates in Agda in Sections \ref{sec:propositions} and
-\ref{sec:predicates}.  We formalize several important concepts in
-§\ref{sec:fun-approx-iter}: functionals, approximation, and
-iteration, and prove some basic results about them.  We construct all
-of the logical connectives of SIL in
-§\ref{sec:open-propositions}, including the definition of
-recursive predicates and their fixpoint theorem.  The proof theory of
-SIL is developed in §\ref{sec:proofs}, where we construct the
-proof rules of this modal logic.
-
-\input{review-appel.tex}
+The primary challenge in constructing the Step-Indexed Logic is in
+proving
 
@@
 
+\input{../../src/latex/EquivalenceRelation.tex}
+
 \input{latex/July2023/STLCDeterministic2.tex}

### 2023-09-26 papers/July2023/paper.tex
check

@@
 
-Unfortunately, the \textsf{fixpoint}ᵒ theorem that we introduced in
-\S~\ref{sec:formula-equivalence} cannot be proved for arbitrary
-functions from ℕ to \textsf{Prop}. As usual, one needs to place some
-restrictions on the function space. We place three restrictions on
-these functions: (1) they must be downward closed
-
-The primary challenge in constructing the Step-Indexed Logic is in
-proving
-
 \input{../../src/latex/StepIndexedLogic2.tex}
